---
title: "Scenario 01: SEIRH + all-or-nothing vaccination"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_01")
baseline <- "scenario_00"
full <- bench$full
base <- bench$every[bench$every$scenario == baseline, ]
attack <- function(data, n) if (nrow(data)) sprintf("%.2g", median(data$attack_rate[data$n == n])) else "?"
```

[Back to the project overview](../README.md) ·
[Scenario 00 report](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md) plus a vaccine given before
transmission starts. The disease, network, seeds, run length, and calibration
are identical. So for any engine, size, and replicate, the two scenarios
differ only in the vaccine, and this report compares them replicate by
replicate.

## Model

### Vaccine

| Parameter (`scenario.toml`) | Value | Meaning                                         |
|:----------------------------|------:|:------------------------------------------------|
| `vaccine_coverage`          |  0.30 | Share of all agents vaccinated                  |
| `vaccine_efficacy`          |  0.80 | Probability that a vaccinated agent is protected |

- Before the first day, `round(vaccine_coverage * n)` agents are chosen
  uniformly at random and vaccinated.
- The vaccine is **all-or-nothing**. Each vaccinated agent is independently
  protected with probability `vaccine_efficacy`. A protected agent can never
  be infected. An unprotected one behaves exactly like an unvaccinated agent.
  A *leaky* vaccine would instead lower everyone's per-contact risk by 80%.
- Vaccination is independent of the initial infections. A seed case that also
  gets vaccinated stays infected.

About 24% of agents end up protected, so the effective reproduction number
falls from about 2 to about 1.5. The disease parameters are those of
scenario 00.

### Outputs

Runners add two fields to each result record:

- `vaccinated`: the number of agents vaccinated.
- `vaccine_protected`: how many of them are protected, including any who were
  also seed cases.

Protected agents who are never infected are counted in `final_susceptible`.
So the five compartments still sum to `n`, and the attack rate is still
`1 - final_susceptible / n`.

### Engine implementations

Each engine uses its own way of expressing the vaccine, so the model-lines
comparison reflects what a user of that engine would write.

- **epiworldR**: a `tool()` with `susceptibility_reduction = 1`. Unprotected
  vaccinees need no tool, so the runner draws the number of protected agents
  from a binomial and lets epiworld place the tool on that many randomly
  chosen agents.
- **epiworld** (C++) and **epiworldpy**: the same tool, placed the same way.
  Each draws the number of protected agents with its own language's random
  numbers, so unlike in scenario 00 the three runners' epidemics differ
  replicate by replicate, though not in distribution. The C++ library also
  has an all-or-nothing `ToolVaccine`, which neither wrapper exposes. It
  decides whether an agent is protected only when the agent is first exposed,
  so the number protected, which every runner reports, is not observable
  with it.
- **Covasim**: two day-0 `cv.simple_vaccine` interventions targeted with
  `subtarget`. One gives the protected agents `rel_sus = 0`, the other leaves
  the remaining vaccinees unchanged. A single `simple_vaccine` would be leaky.
  Covasim tracks `people.vaccinated` itself.
- **EoN**: protected agents start in a status `"V"` that has no transitions,
  so no induced transmission rate ever reaches them.
- **epydemic**: a `V` compartment with no events. It sits outside the S–I
  edge locus that drives infection. The vaccine is drawn in
  `initialCompartments`, which runs inside the timed simulation call; this
  adds well under a millisecond.
- **ixa**: a `VaccineStatus` property (`Unvaccinated`, `Protected`,
  `Unprotected`) on `Person`. The daily step skips protected neighbours when
  it looks for susceptible contacts.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 12
#| fig-height: 5
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### The epiworld family

The wrappers compared with the C++ runner. Here the three draw the number of
protected agents with different random numbers, so they are matched by seed
but do not run identical epidemics. As in
[scenario 00](../scenario_00/README.md#the-language-layer-costs-nothing-measurable-in-the-simulation),
the C++ runner's slower simulation at 100,000 agents comes from where its
arrays land in memory, not from the language layer.

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate, peak hospitalization load, and the
share of agents the vaccine protected.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The calibrated outcomes are closely aligned within each population size"))
}
```

## Cost of the added complexity

### Run time compared with scenario 00

Each replicate is paired with the scenario 00 replicate that has the same
engine, population size, and seed. Values above one mean that this scenario
took longer.

```{r}
#| label: time-versus-baseline
table_time_versus(bench, baseline)
```

Run time does not isolate the cost of the vaccine machinery. The vaccine cuts
the median attack rate from about `r attack(base, 10000)` to
`r attack(full, 10000)` at 10,000 agents and from `r attack(base, 100000)` to
`r attack(full, 100000)` at 100,000. Engines whose work tracks the number of
active infections (EoN, epydemic, the epiworld family, and ixa) get faster for that
reason alone. Covasim's vectorized daily update touches every agent
regardless of the outbreak's size, so its time barely moves. A ratio below
one therefore does not mean the vaccine is free. It means that the engine's
handling of the vaccine costs less than the smaller outbreak saves. A ratio
above one would mean the feature itself is expensive, as epiworldR's
`distribute_tool_to_set()` was before version 0.16.1 (see below).

### Code required to define the model

A rough measure of effort: how much code each engine needs to express this
scenario, and how much it added to scenario 00. The counting rules are
described in the [project overview](../README.md#measuring-implementation-effort),
and the counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench, baseline)
```

EoN and epydemic only need a compartment with no transitions. Covasim has a
vaccine intervention, but it is leaky, so an all-or-nothing vaccine takes two
targeted doses. epiworldR has a tool that removes susceptibility. ixa adds an
agent property that its hand-written transmission step has to consult.

## Interpretation

### An implementation choice that mattered: epiworldR's `distribute_tool_to_set()`

The most direct epiworldR implementation draws the protected agents in R and
hands them to the tool with `distribute_tool_to_set()`. In epiworldR 0.15.1.0
that function was quadratic in the number of agents. Every per-agent
`add_tool()` cloned the tool, and the clone carried a copy of the whole list
of target IDs. Placing the tool on about 24,000 named agents took roughly
4.6 seconds per run at 100,000 agents, against 0.03 seconds with the approach
the runner uses: draw the number of protected agents, and let epiworld place
the tool at random. Both give the same vaccine statistically. epiworldR
0.16.1 fixed the quadratic cost, so either approach is now fast; the runner
keeps the second.

### Calibration

The transmission multipliers are copied from scenario 00. They correct each
engine's transmission semantics rather than anything specific to this
scenario, so they were not recalibrated. The table above shows that the
engines' attack rates stay aligned with the vaccine in place.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
