---
title: "Scenario 01: SEIRH with an all-or-nothing vaccine"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_01")
```

[Back to the project overview](../README.md) ·
[Methods](../docs/methods.md) · [Results](../docs/results.md) ·
[Scenario 00](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md) plus a vaccine given before
transmission starts. The disease, network, seeds, run length, and calibration
factors are scenario 00's, so for any engine, size, and replicate the two
scenarios differ only in the vaccine; the [results](../docs/results.md#cost-of-added-features)
compare them replicate by replicate.

## Model

```{r}
#| label: parameters
table_parameters(bench)
```

- Before the first day, `round(vaccine_coverage * n)` agents are chosen
  uniformly at random and vaccinated, independently of the seed cases.
- The vaccine is **all-or-nothing**: each vaccinee is protected with
  probability `vaccine_efficacy`, and a protected agent can never be
  infected. An unprotected vaccinee behaves like an unvaccinated agent, and a
  seed case that is also vaccinated stays infected. About 24% of agents end up
  protected, so the effective reproduction number falls from about 2 to
  about 1.5.
- Distributing the vaccine is timed as simulation, as seeding is
  ([Timing](../docs/methods.md#timing)).
- Records add `vaccinated` and `vaccine_protected`. Protected agents who are
  never infected count as susceptible, so the attack rate is still
  `1 - final_susceptible / n`.

## Engine notes

Each engine expresses the vaccine with its own tools:

- **The epiworld family**: a tool with `susceptibility_reduction = 1`. Each
  runner draws the number of protected agents (one binomial draw, with its
  own language's random numbers) and lets epiworld place the tool at random,
  so the three runners' epidemics now differ replicate by replicate, though
  not in distribution. epiworld's `ToolVaccine` would decide protection only
  at first exposure, so the number protected would not be observable.
- **Covasim**: two day-0 `cv.simple_vaccine` interventions targeted with
  `subtarget`, one setting `rel_sus = 0` for the protected agents; a single
  one would be leaky.
- **Starsim**: its all-or-nothing `ss.simple_vx(leaky=False)`, delivered by a
  day-0 `ss.campaign_vx`. The runner seeds NumPy's global generator, which
  `simple_vx` draws from.
- **EoN** and **epydemic**: a `V` status or compartment with no transitions.
- **ixa**: a `VaccineStatus` property set in a plan at time 0, which the daily
  step consults. **Agents.jl**: a `protected` field the daily step consults.
- **individual**: a `Bitset` of protected agents that the infection process
  leaves out.
- **FRED**: a second condition, `VAX`, written as in FRED's own vaccine
  example. An import on day 0 moves the vaccinees into it, and `Protected`
  sets their susceptibility to zero.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

[Timing](../docs/methods.md#timing) defines build and simulate.

```{r}
#| label: phases
table_phases(bench)
```

### Memory

Median [Q1, Q3] in MiB ([Memory](../docs/methods.md#memory)).

```{r}
#| label: memory
table_memory(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

### Model lines

Counted as described in
[Implementation size](../docs/methods.md#implementation-size); the last
column is what the vaccine added to scenario 00.

```{r}
#| label: code-effort
table_code_effort(bench, baseline = "scenario_00")
```

## Notes

**The C++ runner is the slowest epiworld runner here at 100,000 agents**,
although it runs the same library. Placing the vaccine allocates about
24,000 tool objects inside `run()`. The C++ runner starts with a small heap,
which grows during those allocations; R and Python grew a large heap while
starting up. Telling glibc to keep a large heap in reserve
(`MALLOC_TOP_PAD_`) or to serve large blocks from the heap
(`MALLOC_MMAP_THRESHOLD_`) brought a median of 10.9 ms down to about 9 ms.
The runner is timed as a C++ user would write it.

**epiworldR's `distribute_tool_to_set()`** was quadratic before 0.16.1: every
per-agent `add_tool()` cloned the tool with a copy of the whole target list,
so placing it on about 24,000 named agents took 4.6 seconds at 100,000
agents, against 0.03 seconds for the approach the runner uses (draw the
count, let epiworld place the tool). Both approaches are fast now.

**Agents.jl's first runner** drew the vaccine in a loop at the script's top
level, over Julia's untyped globals, which doubled its simulation time at
10,000 agents. Moving the loop into a function that the warm-up compiles, as
a Julia user would, removed the cost.
