---
title: "Scenario 01: SEIRH + all-or-nothing vaccination"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_01")
baseline <- "scenario_00"
full <- bench$full
base <- bench$every[bench$every$scenario == baseline, ]
attack <- function(data, n) if (nrow(data)) sprintf("%.2g", median(data$attack_rate[data$n == n])) else "?"
```

[Back to the project overview](../README.md) ·
[Scenario 00 report](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md) plus a vaccine given before
transmission starts. The disease, network, seeds, run length, and calibration
are identical. So for any engine, size, and replicate, the two scenarios
differ only in the vaccine, and this report compares them replicate by
replicate.

## Model

### Vaccine

| Parameter (`scenario.toml`) | Value | Meaning                                         |
|:----------------------------|------:|:------------------------------------------------|
| `vaccine_coverage`          |  0.30 | Share of all agents vaccinated                  |
| `vaccine_efficacy`          |  0.80 | Probability that a vaccinated agent is protected |

- Before the first day, `round(vaccine_coverage * n)` agents are chosen
  uniformly at random and vaccinated.
- The vaccine is **all-or-nothing**. Each vaccinated agent is independently
  protected with probability `vaccine_efficacy`. A protected agent can never
  be infected. An unprotected one behaves exactly like an unvaccinated agent.
  A *leaky* vaccine would instead lower everyone's per-contact risk by 80%.
- Vaccination is independent of the initial infections. A seed case that also
  gets vaccinated stays infected.
- Distributing the vaccine is timed as part of the simulation in every
  runner, as seeding is (see [scenario 00](../scenario_00/README.md)). The
  draw differs in every run, and epiworld places its tools inside `run()`, so
  the other runners time their draw too. EoN's timer covers building its
  initial status dictionary. Covasim's covers the draw, and Covasim gives the
  doses on day 0 inside `sim.run()`. Starsim draws the vaccinees when its
  day-0 campaign runs, inside `sim.run()`. epydemic draws inside its
  simulation call, and ixa in a plan at time 0, inside `execute()`.
  individual's timer covers its draw. The epiworld runners
  draw only the number of protected agents (one binomial draw) before
  `run()`.

About 24% of agents end up protected, so the effective reproduction number
falls from about 2 to about 1.5. The disease parameters are those of
scenario 00.

### Outputs

Runners add two fields to each result record:

- `vaccinated`: the number of agents vaccinated.
- `vaccine_protected`: how many of them are protected, including any who were
  also seed cases.

Protected agents who are never infected are counted in `final_susceptible`.
So the five compartments still sum to `n`, and the attack rate is still
`1 - final_susceptible / n`.

### Engine implementations

Each engine uses its own way of expressing the vaccine, so the model-lines
comparison reflects what a user of that engine would write.

- **epiworldR**: a `tool()` with `susceptibility_reduction = 1`. Unprotected
  vaccinees need no tool, so the runner draws the number of protected agents
  from a binomial and lets epiworld place the tool on that many randomly
  chosen agents.
- **epiworld** (C++) and **epiworldpy**: the same tool, placed the same way.
  Each draws the number of protected agents with its own language's random
  numbers, so unlike in scenario 00 the three runners' epidemics differ
  replicate by replicate, though not in distribution. The C++ library also
  has an all-or-nothing `ToolVaccine`, which neither wrapper exposes. It
  decides whether an agent is protected only when the agent is first exposed,
  so the number protected, which every runner reports, is not observable
  with it.
- **Covasim**: two day-0 `cv.simple_vaccine` interventions targeted with
  `subtarget`. One gives the protected agents `rel_sus = 0`, the other leaves
  the remaining vaccinees unchanged. A single `simple_vaccine` would be leaky.
  Covasim tracks `people.vaccinated` itself.
- **Starsim**: its own all-or-nothing vaccine, `ss.simple_vx(leaky=False)`,
  delivered by a day-0 `ss.campaign_vx` to the drawn agents (the campaign's
  eligibility function returns them). The vaccine sets `rel_sus = 0` for the
  protected vaccinees; Starsim applies interventions before transmission
  within a step. `simple_vx` draws who is protected from NumPy's global
  generator, so the runner seeds it.
- **EoN**: protected agents start in a status `"V"` that has no transitions,
  so no induced transmission rate ever reaches them.
- **epydemic**: a `V` compartment with no events. It sits outside the S–I
  edge locus that drives infection. The vaccine is drawn in
  `initialCompartments`, which runs inside the timed simulation call.
- **ixa**: a `VaccineStatus` property (`Unvaccinated`, `Protected`,
  `Unprotected`) on `Person`, set by a plan at time 0. The daily step skips
  protected neighbours when it looks for susceptible contacts.
- **individual**: no intervention objects, so the protected agents are a
  `Bitset`, and the infection process intersects the susceptible agents with
  its complement.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

The simulation time above covers what an engine has to redo for every
replicate on the same network (see
[what is measured](../README.md#run-time)). *Build* is the rest of the
setup: turning the edge list into the engine's population and network, which
a user does once per network. *Build + simulate* is the time to a first
result once the edge list is in memory. Reading the edge file is shown but
left out of it, because its cost depends on each language's file parsing
rather than on the engine.

```{r}
#| label: phases
table_phases(bench)
```

### The epiworld family

The wrappers compared with the C++ runner. Here the three draw the number of
protected agents with different random numbers, so they are matched by seed
but do not run identical epidemics. Here the C++ runner is slower than the
wrappers at 100,000 agents, although it runs the same library. This comes
from memory allocation, not from the language layer. Placing the vaccine
allocates about 24,000 tool objects inside `run()`. The C++ runner starts
with a small heap, which grows during those allocations; R and Python have
already grown a large heap while starting up. Run alone, the C++ runner
takes a median 10.9 ms here, and about 9 ms when glibc is told to keep a large
heap in reserve (`MALLOC_TOP_PAD_`) or to serve large blocks from the heap
rather than with `mmap` (`MALLOC_MMAP_THRESHOLD_`). The runner is timed as a
C++ user would write it, so the published numbers keep this effect.

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate, peak hospitalization load, and the
share of agents the vaccine protected.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The calibrated outcomes are closely aligned within each population size"))
}
```

## Cost of the added complexity

### Run time compared with scenario 00

Each replicate is paired with the scenario 00 replicate that has the same
engine, population size, and seed. Values above one mean that this scenario
took longer.

```{r}
#| label: time-versus-baseline
table_time_versus(bench, baseline)
```

Run time does not isolate the cost of the vaccine machinery. The vaccine cuts
the median attack rate from about `r attack(base, 10000)` to
`r attack(full, 10000)` at 10,000 agents and from `r attack(base, 100000)` to
`r attack(full, 100000)` at 100,000. Engines whose work tracks the number of
active infections (EoN, epydemic, the epiworld family, and ixa) get faster for that
reason alone. Covasim's vectorized daily update touches every agent
regardless of the outbreak's size, and Starsim's draws transmission on every
edge every day, so their times barely move. individual's infection process
scans the infectious agents' neighbours but also works on full-population
bitsets every day, so it sits in between. A ratio below
one therefore does not mean the vaccine is free. It means that the engine's
handling of the vaccine costs less than the smaller outbreak saves. A ratio
above one would mean the feature itself is expensive, as epiworldR's
`distribute_tool_to_set()` was before version 0.16.1 (see below).

### Code required to define the model

A rough measure of effort: how much code each engine needs to express this
scenario, and how much it added to scenario 00. The counting rules are
described in the [project overview](../README.md#measuring-implementation-effort),
and the counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench, baseline)
```

EoN and epydemic only need a compartment with no transitions. Covasim has a
vaccine intervention, but it is leaky, so an all-or-nothing vaccine takes two
targeted doses. Starsim's own vaccine product supports all-or-nothing
protection directly, and needs only a campaign to deliver it. epiworldR has a
tool that removes susceptibility. ixa adds an agent property that its
hand-written transmission step has to consult, and individual a set that its
infection process leaves out.

## Interpretation

### An implementation choice that mattered: epiworldR's `distribute_tool_to_set()`

The most direct epiworldR implementation draws the protected agents in R and
hands them to the tool with `distribute_tool_to_set()`. In epiworldR 0.15.1.0
that function was quadratic in the number of agents. Every per-agent
`add_tool()` cloned the tool, and the clone carried a copy of the whole list
of target IDs. Placing the tool on about 24,000 named agents took roughly
4.6 seconds per run at 100,000 agents, against 0.03 seconds with the approach
the runner uses: draw the number of protected agents, and let epiworld place
the tool at random. Both give the same vaccine statistically. epiworldR
0.16.1 fixed the quadratic cost, so either approach is now fast; the runner
keeps the second.

### What the vaccine costs epiworld and ixa

Both place the vaccine inside the timed call, before the first day. epiworld
clones the tool onto the heap for each of the roughly 24,000 protected agents
at 100,000 agents, which takes about 3.2 ms per run; ixa sets a property on
each of the 30,000 vaccinees in a plan at time 0, about 1.7 ms. Because the
vaccine also shrinks the outbreak, this fixed cost is a larger share of
epiworld's time here than in scenario 00. ixa's time per replicate is still
mostly rebuilding its context. [analysis.md](../analysis.md) has the
measurements, and explains the rest of the difference between the two
engines.

### Calibration

The transmission multipliers are copied from scenario 00. They correct each
engine's transmission semantics rather than anything specific to this
scenario, so they were not recalibrated. The table above shows that the
engines' attack rates stay aligned with the vaccine in place.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
