# Using the LSTM model to Find the Best Parameters Used in Generating the SIR Model

# Section 1:

#### implementing LSTM model

installing required packages:

```{r}
library(epiworldR)
library(data.table)
library(tensorflow)
library(keras)
library(parallel)
library(keras3)
library(dplyr)
library(ggplot2)



```

## calling Preparation Function

Now we call the source to prepare the data we want to generate in the next step.

```{r}
# devtools::install_github("UofUEpi/epiworldR")
source("calibration/dataprep.R")
```

# Set parameters

```{r}

N     <- 2e4
n     <- 5000
ndays <- 50
ncores <- 20
```

# Generate parameters (theta)

The `generate_simulation_parameters` function generates a table of parameters (`theta`) for the SIR model simulations.

```{r}
set.seed(1231)
generate_simulation_parameters <- function(N, n) {
  theta <- data.table(
    preval = sample((100:2000) / n, N, TRUE),
    crate  = rgamma(N, 5, 1),  # Mean 10
    ptran  = rbeta(N, 3, 7),   # Mean 3/(3 + 7) = 0.3
    prec   = rbeta(N, 10, 10 * 2 - 10)  # Mean 10 / (10 * 2 - 10) = .5
  )
  theta[, hist(crate)]
  return(theta)
}
```

-   **Inputs**:

    -   `N`: The number of simulations or rows of parameter sets to generate.

    -   `n`: The population size used for prevalence calculations.

-   **Outputs**:

    -   The function returns a `data.table` containing four key epidemiological parameters:

        1.  **`preval`**: Prevalence, or the initial proportion of the population that is infected. It is sampled from a range between `100/n` and `2000/n` (which corresponds to infection rates between 0.02% and 0.4%).

        2.  **`crate`**: Contact rate, which is the rate at which individuals in the population come into contact with others. It is generated from a Gamma distribution with a shape parameter of 5 and a rate parameter of 1 (which gives a mean contact rate of 5).

        3.  **`ptran`**: Transmission probability, or the likelihood of disease transmission upon contact, drawn from a Beta distribution with shape parameters 3 and 7. This gives a mean transmission rate of 0.3 (since 3 / (3 + 7) = 0.3).

        4.  **`prec`**: Recovery probability, or the likelihood of recovery after infection. It is drawn from a Beta distribution with shape parameters of 10 and 10, giving a mean recovery rate of 0.5 (since 10 / (10 \* 2 - 10) = 0.5).

# run epidemic model simulations in parallel

`run_epidemic_simulations` Function runs SIR (Susceptible-Infectious-Recovered) model simulations in parallel for `N` iterations, using a set of parameters (`theta`) for each simulation.

```{r}
run_epidemic_simulations <- function(N, theta, seeds, n, ndays, ncores) {
  matrices <- parallel::mclapply(1:N, FUN = function(i) {
    fn <- sprintf("calibration/simulated_data/sir-%06i.rds", i)
    if (file.exists(fn)) return(readRDS(fn))
    set.seed(seeds[i])
    m <- theta[i, ModelSIRCONN("mycon", prevalence = preval, contact_rate = crate, transmission_rate = ptran, recovery_rate = prec, n = n)]
    verbose_off(m)
    run(m, ndays = ndays)
    ans <- prepare_data(m)
    saveRDS(ans, fn)
    return(ans)
  }, mc.cores = ncores)
  return(matrices)
}
```

```{r}
filter_valid_simulations <- function(matrices, theta) {
  valid_indices <- intersect(
    which(!sapply(matrices, inherits, what = "error")),
    which(!sapply(matrices, \(x) any(is.na(x))))
  )
  matrices <- matrices[valid_indices]
  theta <- theta[valid_indices, ]
  return(list(matrices = matrices, theta = theta, N = length(valid_indices)))
}
```

**Inputs**:

-   `N`: Number of simulations to run.

-   `theta`: A table containing the parameters for each simulation (prevalence, contact rate, transmission rate, and recovery rate).

-   `seeds`: A set of random seeds to ensure each simulation has different random behavior.

-   `n`: Population size for the simulation.

-   `ndays`: Number of days to simulate.

-   `ncores`: Number of CPU cores to use for parallel processing.

# prepare simulation data for neural network training

`prepare_neural_network_data` Function prepares the simulation data and parameters for use in training a neural network.

```{r}
# Function to prepare simulation data for neural network training
prepare_neural_network_data <- function(N, matrices, theta) {
  arrays_1d <- array(dim = c(N, dim(matrices[[1]][1, , ])))
  for (i in seq_along(matrices)) {
    arrays_1d[i, , ] <- matrices[[i]][1, , ]
  }
  theta2 <- copy(theta)
  theta2[, crate := plogis(crate / 10)]
  return(list(arrays_1d = arrays_1d, theta2 = theta2))
}

# Function to save the prepared data to an RDS file
save_prepared_data <- function(theta2, arrays_1d) {
  saveRDS(
    list(
      theta = theta2,
      simulations = arrays_1d
    ),
    file = "calibration/sir.rds",
    compress = TRUE
  )
}
```

# Split data into training and test sets

```{r}
split_train_test_data <- function(N, arrays_1d, theta2) {
  N_train <- floor(N * 0.7)
  id_train <- 1:N_train
  id_test <- (N_train + 1):N
  train <- list(
    x = array_reshape(arrays_1d[id_train, , ], dim = c(N_train, dim(arrays_1d)[-1])),
    y = array_reshape(as.matrix(theta2)[id_train, ], dim = c(N_train, ncol(theta2)))
  )
  test <- list(
    x = array_reshape(arrays_1d[id_test, , ], dim = c(N - N_train, dim(arrays_1d)[-1])),
    y = array_reshape(as.matrix(theta2)[id_test, ], dim = c(N - N_train, ncol(theta2)))
  )
  return(list(train = train, test = test))
}
```

-   **Inputs**:

    -   `N`: The number of valid simulations.

    -   `matrices`: A list of simulation results (from `run_epidemic_simulations`).

    -   `theta`: A table of parameters used for the simulations.

-   **Functionality**:

    -   **Creating an Array for Simulations**:

        -   `arrays_1d`: A 3D array is created to store the first slice of data (e.g., infections) from each simulation result. The dimensions are `(N, rows, cols)` where `N` is the number of simulations.

        -   Each simulation's result is extracted (`matrices[[i]][1, , ]`) and added to `arrays_1d`.

    -   **Adjusting Parameters**:

        -   `theta2`: A copy of `theta` is created, and the contact rate (`crate`) is transformed using the logistic function (`plogis(crate / 10)`) to map values into a probability range (0, 1).

# Build and train the LSTM model

`build_and_train_lstm` Function builds and trains an LSTM (Long Short-Term Memory) neural network model using the `keras3` package.

```{r}
build_and_train_lstm <- function(train, theta2, arrays_1d) {
  model <- keras3::keras_model_sequential() %>%
    keras3::layer_lstm(units = 64, input_shape = c(dim(arrays_1d)[2], dim(arrays_1d)[3]), return_sequences = FALSE) %>%
    keras3::layer_dense(units = ncol(theta2), activation = 'sigmoid')
  
  model %>% compile(optimizer = 'adam', loss = 'mse', metrics = 'accuracy')
  
  tensorflow::set_random_seed(331)
  model %>% fit(train$x, train$y, epochs = 100, verbose = 2)
  
  return(model)
}
```

**Inputs**:

-   `train`: The training dataset, consisting of `train$x` (input data) and `train$y` (target parameters).

-   `theta2`: The adjusted table of parameters that the model will learn to predict.

-   `arrays_1d`: The 3D array containing simulation data, which helps define the input shape of the LSTM.

-   `lstm_units`: The number of LSTM units (default is 64) in the hidden layer.

-   `epochs`: Number of training epochs (default is 100), i.e., how many times the model will iterate over the training data.

-   `loss_function`: The loss function used to optimize the model (default is Mean Squared Error, `'mse'`).

-   `optimizer`: The optimizer for training (default is Adam).

-   `activation`: The activation function used in the output layer (default is `'sigmoid'`).

# predict using the trained model

```{r}
# Function to predict using the trained model
make_predictions <- function(model, test, theta) {
  pred <- predict(model, x = test$x) |> as.data.table() |> setnames(colnames(theta))
  return(pred)
}

# Function to compute the Mean Absolute Error (MAE)
compute_mae <- function(pred, test) {
  MAEs <- abs(pred - as.matrix(test$y)) |> colMeans() |> print()
  return(MAEs)
}
```

# Visualize model predictions against observed data

```{r}
visualize_predictions <- function(pred, test, MAEs, theta, N, N_train) {
  pred[, id := 1L:.N]
  pred[, crate := qlogis(crate) * 10]
  pred_long <- melt(pred, id.vars = "id")
  
  theta_long <- test$y |> as.data.table()
  setnames(theta_long, names(theta))
  theta_long[, id := 1L:.N]
  theta_long[, crate := qlogis(crate) * 10]
  theta_long <- melt(theta_long, id.vars = "id")
  
  alldat <- rbind(cbind(pred_long, Type = "Predicted"), cbind(theta_long, Type = "Observed"))
  
  ggplot(alldat, aes(x = value, colour = Type)) +
    facet_wrap(~variable, scales = "free") +
    geom_boxplot()
  
  alldat_wide <- dcast(alldat, id + variable ~ Type, value.var = "value")
  
  vnames <- data.table(
    variable = c("preval", "crate", "ptran", "prec"),
    Name     = paste(c("Init. state", "Contact Rate", "P(transmit)", "P(recover)"), sprintf("(MAE: %.2f)", MAEs))
  )
  
  alldat_wide <- merge(alldat_wide, vnames, by = "variable")
  
  ggplot(alldat_wide, aes(x = Observed, y = Predicted)) +
    facet_wrap(~ Name, scales = "free") +
    geom_abline(slope = 1, intercept = 0) +
    geom_point(alpha = .2) +
    labs(
      title = "Observed vs Predicted (validation set)",
      subtitle = sprintf("The model includes %i simulated datasets, of which %i were used for training.", N, N_train),
      caption = "Predictions made using a CNN as implemented with loss function MAE."
    )
}
```

# Main execution block

```{r}
theta <- generate_simulation_parameters(N, n)
seeds <- sample.int(.Machine$integer.max, N, TRUE)
matrices <- run_epidemic_simulations(N, theta, seeds, n, ndays, ncores)

# Filter and prepare data
valid_data <- filter_valid_simulations(matrices, theta)
N <- valid_data$N
theta <- valid_data$theta
matrices <- valid_data$matrices

# Prepare for neural network
nn_data <- prepare_neural_network_data(N, matrices, theta)
arrays_1d <- nn_data$arrays_1d
theta2 <- nn_data$theta2

# Save data
save_prepared_data(theta2, arrays_1d)

# Split train and test data
train_test_data <- split_train_test_data(N, arrays_1d, theta2)
train <- train_test_data$train
test <- train_test_data$test
```

```{r,epochs=FALSE}
# Build, train, and evaluate the model
model <- build_and_train_lstm(train, theta2, arrays_1d)
```

```{r}
pred <- make_predictions(model, test, theta)
```

```{r}
MAEs <- compute_mae(pred, test)
print(MAEs)
```

```{r}
# Visualize the results
visualize_predictions(pred, test, MAEs, theta, N, floor(N * 0.7))

```

# Section 2:

# Grid Search function for LSTM hyperparameters

The `grid_search_lstm` function performs a grid search over different hyperparameter combinations for training an LSTM model.

```{r}
grid_search_lstm <- function(train, test, theta2, arrays_1d, lstm_units_options, epochs_options, loss_function_options, optimizer_options, activation_options) {
  results <- list()
  
  for (lstm_units in lstm_units_options) {
    for (epochs in epochs_options) {
      for (loss_function in loss_function_options) {
        for (optimizer in optimizer_options) {
          for (activation in activation_options) {
            
            # Train the model with current hyperparameters
            model <- build_and_train_lstm(train, theta2, arrays_1d, lstm_units, epochs, loss_function, optimizer, activation)
            mae <- evaluate_model(model, test)
            
            # Store the results
            results[[length(results) + 1]] <- list(
              lstm_units = lstm_units, 
              epochs = epochs, 
              loss_function = loss_function,
              optimizer = optimizer, 
              activation = activation, 
              mae = mae, 
              model = model
            )
          }
        }
      }
    }
  }
  return(results)
}


```

**Inputs**:

-   `train`, `test`: The training and testing datasets.

-   `theta2`, `arrays_1d`: Additional parameters that might influence the model or input data (specific to the problem).

-   `lstm_units_options`, `epochs_options`, `loss_function_options`, `optimizer_options`, `activation_options`: These are lists of possible values for each hyperparameter, which will be used in the grid search (LSTM units, epochs, loss functions, optimizers, and activation functions).

# Find the best model based on MAE

```{r}
find_best_model <- function(results) {
  best_result <- results[[which.min(sapply(results, function(x) x$mae))]]
  return(best_result)
}


```

# Main function to run everything and find the best model

This `run_lstm_grid_search` function performs a grid search over a set of hyperparameters for an LSTM model to find the best-performing configuration based on Mean Absolute Error (MAE).

```{r,epochs=FALSE}
run_lstm_grid_search <- function(train, test, theta2, arrays_1d) {
  
  # Define the hyperparameter grid
  lstm_units_options <- c(64, 128, 256)
  epochs_options <- c(10)
  loss_function_options <- c('mse', 'mae')
  optimizer_options <- c('adam', 'rmsprop')
  activation_options <- c('sigmoid', 'relu')
  
  # Run grid search
  results <- grid_search_lstm(train, test, theta2, arrays_1d, lstm_units_options, epochs_options, loss_function_options, optimizer_options, activation_options)
  
  # Find the best model based on MAE
  best_result <- find_best_model(results)
  
  # Print the best configuration and MAE
  cat("Best model configuration:\n")
  print(best_result)
  
  return(best_result)
}

# Assuming train, test, theta2, and arrays_1d are already defined
best_model_result <- run_lstm_grid_search(train, test, theta2, arrays_1d)
```

**Define Hyperparameter Grid**:

-   Lists of possible values for LSTM units, epochs, loss functions, optimizers, and activation functions are defined:

    -   `lstm_units_options`: Number of units in the LSTM layer (64, 128, 256).

    -   `epochs_options`: Number of training epochs (10).

    -   `loss_function_options`: Loss functions ('mse' and 'mae').

    -   `optimizer_options`: Optimizers ('adam' and 'rmsprop').

    -   `activation_options`: Activation functions ('sigmoid' and 'relu').

##### See the history of the best model

```{r}
best_model_result$model
```

# Run the best model and visualization

```{r}
# model <- best_model_result$model
build_and_train_lstm <- function(train, theta2, arrays_1d) {
  model <- keras3::keras_model_sequential() %>%
    keras3::layer_lstm(units = 256, input_shape = c(dim(arrays_1d)[2], dim(arrays_1d)[3]), return_sequences = FALSE) %>%
    keras3::layer_dense(units = ncol(theta2), activation = 'sigmoid')
  
  model %>% compile(optimizer = 'adam', loss = 'mse', metrics = 'accuracy')
  
  tensorflow::set_random_seed(331)
  model %>% fit(train$x, train$y, epochs = 100, verbose = 2)
  
  return(model)
}
model=build_and_train_lstm(train, theta2, arrays_1d)
model %>% fit(
  train$x,
  train$y,
  epochs = 100,
  verbose = 2
  )

pred <- predict(model, x = test$x) |>
  as.data.table() |>
  setnames(colnames(theta))

MAEs <- abs(pred - as.matrix(test$y)) |>
  colMeans() |>
  print()

# save_model_hdf5(model, "sir-keras")

# Visualizing ------------------------------------------------------------------
pred[, id := 1L:.N]
pred[, crate := qlogis(crate) * 10]
pred_long <- melt(pred, id.vars = "id")

theta_long <- test$y |> as.data.table()
setnames(theta_long, names(theta))
theta_long[, id := 1L:.N]
theta_long[, crate := qlogis(crate) * 10]
theta_long <- melt(theta_long, id.vars = "id")

alldat <- rbind(
  cbind(pred_long, Type = "Predicted"),
  cbind(theta_long, Type = "Observed")
)

library(ggplot2)

```

```         
```

```{r}
ggplot(alldat, aes(x = value, colour = Type)) +
  facet_wrap(~variable, scales = "free") +
  geom_boxplot()


```

```{r}
alldat_wide <- dcast(alldat, id + variable ~ Type, value.var = "value")

vnames <- data.table(
  variable = c("preval", "crate", "ptran", "prec"),
  Name     = paste(
    c("Init. state", "Contact Rate", "P(transmit)", "P(recover)"),
    sprintf("(MAE: %.2f)", MAEs)
    )
)

alldat_wide <- merge(alldat_wide, vnames, by = "variable")
```

```{r}
ggplot(alldat_wide, aes(x = Observed, y = Predicted)) +
  facet_wrap(~ Name, scales = "free") +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(alpha = .2) +
  labs(
    title    = "Observed vs Predicted (validation set)",
    subtitle = sprintf(
      "The model includes %i simulated datasets, of which %i were used for training.",
      N,
      N_train
      ),
    caption  = "Predictions made using a LSTM as implemented with loss function MAE."
    
    )


```
