---
title: "Network epidemic ABM speed benchmark"
subtitle: "epiworldR, Covasim, EoN, and epydemic"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup
options(scipen = 999)
results_path <- file.path("results", "results.csv")
manifest_path <- file.path("results", "run-manifest.json")
if (file.exists(results_path)) {
  all_results <- read.csv(results_path, stringsAsFactors = FALSE)
} else {
  all_results <- data.frame()
}
manifest <- if (file.exists(manifest_path)) jsonlite::read_json(manifest_path, simplifyVector = TRUE) else NULL

if (nrow(all_results)) {
  full <- subset(all_results, days == 100 & n %in% c(10000, 100000))
} else {
  full <- all_results
}

has_full <- nrow(full) > 0
expected_per_cell <- 100L
complete_design <- has_full && all(
  table(factor(full$engine, levels = c("epiworldR", "covasim", "EoN", "epydemic")),
        factor(full$n, levels = c(10000, 100000))) >= expected_per_cell
)
```

## Executive summary

This benchmark compares simulation speed for 100-day network epidemics at
10,000 and 100,000 agents, with 100 independent replicates per engine and size. The included engines are [epiworldR](https://uofuepibio.github.io/epiworldR/){target="_blank"} [@meyerEpiworldRFastAgentBased2023], [Covasim](https://covasim.org/){target="_blank"} [@kerrCovasimAgentbasedModel2021], [EoN](https://epidemicsonnetworks.readthedocs.io/en/latest/EoN.html){target="_blank"} [@millerEoNEpidemicsNetworks2019], and [epydemic](https://github.com/simoninireland/epydemic){target="_blank"} [@dobson2022epydemic]. 
All engines use the exact same cached Watts--Strogatz edge list for a population
size (mean degree 10, rewiring probability 0.05) and target an early-outbreak
$R_0 = 2$. The epiworldR cells apply size-specific transmission multipliers of
0.887 at 10,000 agents and 0.855 at 100,000 agents. The restricted Covasim
cells apply corresponding factors of 0.715 and 0.710. These empirical factors
align realized attack rates under the frameworks' different transmission
semantics.

```{r}
#| label: completion-status
#| results: asis
if (!has_full) {
  cat("::: {.callout-warning}\n",
      "No full-profile results are present yet. Run `make benchmark`, then render this report again. ",
      "The available smoke results validate integration but are not performance evidence.\n",
      ":::\n", sep = "")
} else if (!complete_design) {
  cat("::: {.callout-warning}\n",
      "The full-profile cache is incomplete. Estimates below are provisional; rerun `make benchmark` to resume missing cells.\n",
      ":::\n", sep = "")
} else {
  cat("::: {.callout-tip}\nThe complete 800-run design is available.\n:::\n")
}
```

```{r}
#| label: environment
if (!is.null(manifest)) {
  environment <- data.frame(
    Field = c("Platform", "Python", "Workers", "Latest run failures"),
    Value = c(manifest$platform, manifest$python, manifest$workers, manifest$failures)
  )
  knitr::kable(environment, row.names = FALSE)
}
```

```{r}
#| label: summary-table
if (has_full) {
  q <- function(x, p) unname(stats::quantile(x, p, na.rm = TRUE))
  split_rows <- split(full, list(full$engine, full$n), drop = TRUE)
  summary_rows <- do.call(rbind, lapply(split_rows, function(d) data.frame(
    engine = d$engine[[1]],
    agents = d$n[[1]],
    replicates = nrow(d),
    median_sim_s = median(d$simulate_seconds),
    q25_sim_s = q(d$simulate_seconds, 0.25),
    q75_sim_s = q(d$simulate_seconds, 0.75)
  )))
  summary_rows <- summary_rows[order(summary_rows$agents, summary_rows$median_sim_s), ]
  knitr::kable(summary_rows, digits = 3,
    col.names = c("Engine", "Agents", "Runs", "Median simulation (s)",
                  "Q1 (s)", "Q3 (s)"),
    row.names = FALSE)
}
```

## Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 9
#| fig-height: 5
if (has_full) {
  full$engine <- factor(full$engine, levels = c("epiworldR", "covasim", "EoN", "epydemic"))
  full$agents <- factor(full$n, levels = c(10000, 100000),
                        labels = c("10,000 agents", "100,000 agents"))
  print(
    ggplot2::ggplot(full, ggplot2::aes(engine, simulate_seconds, fill = engine)) +
      ggplot2::geom_boxplot(width = 0.65, outlier.alpha = 0.25) +
      ggplot2::facet_wrap(~ agents, scales = "free_y") +
      ggplot2::scale_y_log10() +
      ggplot2::labs(x = NULL, y = "Simulation wall time\n(seconds, log scale)",
                    subtitle = "One box per 100-day population-size cell") +
      ggplot2::theme_minimal(base_size = 18) +
      ggplot2::theme(
        legend.position = "none",
        axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
        )
  )
}
```

## Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
if (has_full && any(full$engine == "epiworldR")) {
  baseline <- full[full$engine == "epiworldR", c("n", "seed", "simulate_seconds")]
  names(baseline)[3] <- "epiworld_seconds"
  ratios <- merge(full[full$engine != "epiworldR", ], baseline, by = c("n", "seed"))
  ratios$speed_ratio <- ratios$simulate_seconds / ratios$epiworld_seconds
  ratio_groups <- split(ratios, list(ratios$engine, ratios$n), drop = TRUE)
  ratio_table <- do.call(rbind, lapply(ratio_groups, function(d) data.frame(
    engine = as.character(d$engine[[1]]),
    agents = d$n[[1]],
    median = median(d$speed_ratio),
    q25 = unname(quantile(d$speed_ratio, .25)),
    q75 = unname(quantile(d$speed_ratio, .75))
  )))
  knitr::kable(ratio_table, digits = 2,
    col.names = c("Engine", "Agents", "Median time / epiworldR", "Q1", "Q3"),
    row.names = FALSE)
}
```

## Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show final attack rate and peak hospitalization load; differences
also expose the engines' non-identical time semantics and Covasim's native
symptom/severe bookkeeping.

```{r}
#| label: outcomes-table
if (has_full) {
  full$attack_rate <- 1 - full$final_susceptible / full$n
  outcomes <- aggregate(cbind(attack_rate, peak_hospitalized) ~ engine + n,
                        full, median)
  knitr::kable(outcomes, digits = 3,
    col.names = c("Engine", "Agents", "Median final attack rate",
                  "Median peak hospitalized"), row.names = FALSE)
}
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (has_full) {
  outcomes$agents <- factor(outcomes$n, levels = c(10000, 100000),
                             labels = c("10,000", "100,000"))
  print(
    ggplot2::ggplot(outcomes, ggplot2::aes(engine, attack_rate, fill = engine)) +
      ggplot2::geom_col(width = 0.65) +
      ggplot2::facet_wrap(~ agents) +
      ggplot2::scale_y_continuous(limits = c(0, NA)) +
      ggplot2::labs(x = NULL, y = "Median final attack rate",
                    subtitle = "The calibrated outcomes are closely aligned within each population size") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::theme(legend.position = "none")
  )
}
```

## Design and interpretation

The common disease graph is susceptible $\rightarrow$ exposed $\rightarrow$
infectious $\rightarrow$ recovered, with a competing infectious $\rightarrow$
hospitalized $\rightarrow$ recovered branch. Mean latent and infectious
periods are 4 and 7 days, lifetime hospitalization probability is 5%, and the
hospital stay is 7 days. Initial infections are 100 in the full profile.

The shared analytic transmission mapping produced different realized attack
rates across frameworks. Calibration therefore selected size-specific factors
for epiworldR (0.887 at 10,000 agents and 0.855 at 100,000) and for restricted
Covasim (0.715 and 0.710). All 100 replicates in each adjusted cell were then
rerun. These empirical corrections are intentionally population-specific; they
align benchmark outcomes but mean that neither calibrated engine has a strictly
analytic $R_0$ of 2. The values live in `config.toml` and participate in the
cache fingerprint.

epiworldR and epydemic use synchronous daily transitions. EoN uses
continuous-time Gillespie hazards. Covasim retains its native exposed,
infectious, symptomatic, and severe bookkeeping, with severe prevalence used
as the hospitalization proxy. Its runner disables waning immunity and fixes
individual transmissibility and viral load to one, making recovered people
permanently removed and removing those sources of heterogeneity. Covasim does
not expose a public switch to remove the remaining symptom/severity
bookkeeping. Thus this report measures restricted, representative framework
throughput under aligned network and disease targets; it does not claim
bit-for-bit epidemiological equivalence.

The sparse contact network fixes mean degree rather than literal graph density.
At 10,000 and 100,000 agents its densities are approximately 0.001 and 0.0001,
respectively, while every agent still has about ten contacts. This prevents the
edge count, runtime, and memory from growing quadratically.

More details regarding the project setup can be found in [setup.md](./setup.md).

## Reproducibility and cache

Every successful replicate is an atomic JSON cache record keyed by model
configuration, runner source, engine version, and contact-network SHA-256.
Rerunning `make benchmark` schedules only missing or stale records. The default
worker count is one, and common native math-library thread counts are pinned to
one; users must explicitly opt into a second worker.

```{r}
#| label: versions
if (nrow(all_results)) {
  versions <- unique(all_results[c("engine", "engine_version")])
  knitr::kable(versions[order(versions$engine), ],
               col.names = c("Engine", "Recorded version"), row.names = FALSE)
}
```

## References
