---
title: "Network epidemic ABM speed benchmark"
subtitle: "epiworld (C++, R, and Python), Covasim, Starsim, EoN, epydemic, ixa, individual, FRED, and Agents.jl"
format:
  gfm:
    toc: true
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("report/common.R")
all <- load_all(".")
```

This project compares how fast eleven epidemic simulation engines run the same
agent-based model on the same contact network, and how much code each engine
needs to express that model. Three of them are the epiworld family: the
header-only C++ library
[epiworld](https://github.com/UofUEpiBio/epiworld){target="_blank"} [@vegayonEpiworld2026]
and its two wrappers,
[epiworldR](https://uofuepibio.github.io/epiworldR/){target="_blank"} [@meyerEpiworldRFastAgentBased2023]
for R and
[epiworldpy](https://github.com/UofUEpiBio/epiworldpy){target="_blank"} [@vegayonEpiworldpy2026]
for Python. The other eight are
[Covasim](https://covasim.org/){target="_blank"} [@kerrCovasimAgentbasedModel2021]
and its successor
[Starsim](https://starsim.org/){target="_blank"} [@kerrStarsim2025],
[EoN](https://epidemicsonnetworks.readthedocs.io/en/latest/EoN.html){target="_blank"} [@millerEoNEpidemicsNetworks2019],
[epydemic](https://github.com/simoninireland/epydemic){target="_blank"} [@dobson2022epydemic],
[ixa](https://ixa.rs/){target="_blank"} [@ixa2026], the R package
[individual](https://mrc-ide.github.io/individual/){target="_blank"} [@charlesIndividual2021],
[FRED](https://github.com/PublicHealthDynamicsLab/FRED){target="_blank"} [@grefenstetteFRED2013],
and the Julia package
[Agents.jl](https://github.com/JuliaDynamics/Agents.jl){target="_blank"} [@datserisAgentsjl2024].

The benchmark is organized as **scenarios** of increasing complexity. Each
scenario adds a feature to an earlier one, so together they show how run time,
memory, and implementation size grow as the model grows. Every engine runs the
same SEIRH model on the same contact network within each scenario, one
replicate at a time, inside one container image.

- [**Methods**](docs/methods.md): the engines, the common model, networks,
  calibration, the execution protocol, and what is measured.
- [**Engines**](docs/engines.md): what each of the engines is, what it was built
  for, and how it represents a model.
- [**Results**](docs/results.md): run time, scaling, the cost of each added
  feature, memory, epidemiological agreement, and implementation size, across
  every scenario.
- [**Showcase models**](docs/showcase-models.md): the plan for adding, for
  each engine, a scenario it was built for.
- [**Running the benchmark**](setup.md): the container, commands, and the
  published records. [scenarios.md](scenarios.md) explains how to add a
  scenario.

## Scenarios

| Scenario | Model | Report |
|:---------|:------|:-------|
| 00 | SEIRH epidemic with a hospitalization branch, no interventions | [scenario_00/README.md](scenario_00/README.md) |
| 01 | Scenario 00 plus a day-0 all-or-nothing vaccine (30% coverage, 80% efficacy) | [scenario_01/README.md](scenario_01/README.md) |
| 02 | Scenario 01 plus four outputs: transmission tree, daily incidence, reproductive number, and transition matrix | [scenario_02/README.md](scenario_02/README.md) |
| 03 | Scenario 00 at 1,000,000 agents | [scenario_03/README.md](scenario_03/README.md) |
| 04 | Scenario 00 on a real contact network collapsed from [GeoPops](https://github.com/GeoPopsHub){target="_blank"} | [scenario_04/README.md](scenario_04/README.md) |

## Summary

Medians across replicates, with each scenario at its largest population size
(in parentheses). Scenarios 00 to 02 also run at 10,000 agents; the
[results](docs/results.md) and each scenario's report show every size,
interquartile ranges, and the other phases of a run. Engines are ordered from
fastest (or smallest) overall at the top; the axes are logarithmic, because
the engines differ by orders of magnitude.

**Simulation time**: everything an engine redoes for each replicate on the
same network ([Timing](docs/methods.md#timing)).

```{r}
#| label: summary-time-plot
#| fig-width: 11
#| fig-height: 4.2
#| fig-cap: "Median simulation time per replicate (seconds, log scale), by engine and scenario."
plot_summary_time(all)
```

<details>
<summary>The same medians as a table (seconds)</summary>

```{r}
#| label: summary-time
table_summary_time(all)
```

</details>

**Memory**: the overall peak resident memory of a replicate's process
([Memory](docs/methods.md#memory)).

```{r}
#| label: summary-memory-plot
#| fig-width: 11
#| fig-height: 4.2
#| fig-cap: "Median overall peak memory per replicate (MiB, log scale), by engine and scenario."
plot_summary_memory(all)
```

<details>
<summary>The same medians as a table (MiB)</summary>

```{r}
#| label: summary-memory
table_summary_memory(all)
```

</details>

## Running the benchmark

The only host requirement is podman (or Docker, with `CONTAINER=docker`):

```sh
make container-image
make container-check
make container-smoke
make container-benchmark     # every scenario; SCENARIOS=scenario_01 for a subset
make container-report        # this overview, docs/, and every scenario report
```

## Repository layout

```
config.toml          shared study, network, resource, and smoke settings
run.py, scripts/     orchestration, network generation, environment records
report/common.R      tables and figures shared by every report
docs/
  methods.qmd        how the benchmark is built and run (rendered to .md)
  results.qmd        results across scenarios (rendered to .md)
  epiworld-ixa.md    where epiworld's and ixa's time goes
  engines.md         what each engine is (rendered from engines.qmd)
  showcase-models.md the plan for a scenario per engine
scenario_NN/
  README.qmd         the scenario's report (rendered to README.md)
  scenario.toml      model parameters and calibration
  code_regions.yml   regions counted as model code
  runners/           one runner per engine
results/
  runs/              one record per replicate (the source of every table)
  environments/      where each scenario's latest run ran
  results.csv, daily.csv, scenarios.json   derived from runs/
analysis/            the side measurements behind docs/epiworld-ixa.md
```

## References
