---
title: "Network epidemic ABM speed benchmark"
subtitle: "epiworldR, Covasim, EoN, epydemic, and ixa"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup
options(scipen = 999, knitr.kable.NA = "")
results_path <- file.path("results", "results.csv")
manifest_path <- file.path("results", "run-manifest.json")
scenarios_path <- file.path("results", "scenarios.json")
if (file.exists(results_path)) {
  all_results <- read.csv(results_path, stringsAsFactors = FALSE)
} else {
  all_results <- data.frame()
}
engine_levels <- c("epiworldR", "covasim", "EoN", "epydemic", "ixa")
size_levels <- c(10000, 100000)
manifest <- if (file.exists(manifest_path)) jsonlite::read_json(manifest_path, simplifyVector = TRUE) else NULL
scenario_info <- if (file.exists(scenarios_path)) jsonlite::read_json(scenarios_path) else list()
scenario_levels <- sort(names(scenario_info))
baseline <- "scenario_00"
scenario_label <- function(id) {
  vapply(id, function(s) sprintf("%s: %s", sub("scenario_", "", s), scenario_info[[s]]$name), "")
}

if (nrow(all_results)) {
  full <- subset(all_results, days == 100 & n %in% size_levels & scenario %in% scenario_levels)
} else {
  full <- all_results
}

has_full <- nrow(full) > 0
expected_per_cell <- 100L
design_runs <- length(scenario_levels) * length(engine_levels) * length(size_levels) * expected_per_cell
complete_design <- has_full && all(
  table(factor(full$scenario, levels = scenario_levels),
        factor(full$engine, levels = engine_levels),
        factor(full$n, levels = size_levels)) >= expected_per_cell
)
if (has_full) {
  full$scenario_label <- factor(scenario_label(full$scenario), levels = scenario_label(scenario_levels))
}
q <- function(x, p) unname(stats::quantile(x, p, na.rm = TRUE))
```

## Executive summary

This benchmark compares simulation speed for 100-day network epidemics at
10,000 and 100,000 agents, with 100 independent replicates per engine, size,
and scenario. The included engines are [epiworldR](https://uofuepibio.github.io/epiworldR/){target="_blank"} [@meyerEpiworldRFastAgentBased2023], [Covasim](https://covasim.org/){target="_blank"} [@kerrCovasimAgentbasedModel2021], [EoN](https://epidemicsonnetworks.readthedocs.io/en/latest/EoN.html){target="_blank"} [@millerEoNEpidemicsNetworks2019], [epydemic](https://github.com/simoninireland/epydemic){target="_blank"} [@dobson2022epydemic], and [ixa](https://ixa.rs/){target="_blank"} [@ixa2026].
All engines use the exact same cached Watts--Strogatz edge list for a population
size (mean degree 10, rewiring probability 0.05) and target an early-outbreak
$R_0 = 2$. epiworldR and EoN run at the analytic transmission rate. The
restricted Covasim cells apply size-specific transmission multipliers of 0.715
at 10,000 agents and 0.710 at 100,000 agents, and the ixa cells apply 0.983
and 0.988. These empirical factors
align realized attack rates under the frameworks' different transmission
semantics.

The benchmark is organized in scenarios of increasing complexity, so that it
shows how run time and the code needed to define the model change as the
model grows. Each scenario is a folder with its own specification and one
runner per engine; [scenarios.md](scenarios.md) describes how to add one.

```{r}
#| label: scenarios
if (length(scenario_info)) {
  knitr::kable(data.frame(
    Scenario = sub("scenario_", "", scenario_levels),
    Name = vapply(scenario_levels, function(s) scenario_info[[s]]$name, ""),
    Description = vapply(scenario_levels, function(s) scenario_info[[s]]$description, ""),
    Details = sprintf("[%s/README.md](%s/README.md)", scenario_levels, scenario_levels)
  ), row.names = FALSE)
}
```

```{r}
#| label: completion-status
#| results: asis
if (!has_full) {
  cat("::: {.callout-warning}\n",
      "No full-profile results are present yet. Run `make benchmark`, then render this report again. ",
      "The available smoke results validate integration but are not performance evidence.\n",
      ":::\n", sep = "")
} else if (!complete_design) {
  cat("::: {.callout-warning}\n",
      "The full-profile cache is incomplete. Estimates below are provisional; rerun `make benchmark` to resume missing cells.\n",
      ":::\n", sep = "")
} else {
  cat("::: {.callout-tip}\nThe complete ", format(design_runs, big.mark = ","),
      "-run design is available.\n:::\n", sep = "")
}
```

```{r}
#| label: environment
if (!is.null(manifest)) {
  environment <- data.frame(
    Field = c("Platform", "Python", "Workers", "Latest run failures"),
    Value = c(manifest$platform, manifest$python, manifest$workers, manifest$failures)
  )
  knitr::kable(environment, row.names = FALSE)
}
```

```{r}
#| label: summary-table
if (has_full) {
  split_rows <- split(full, list(full$scenario, full$engine, full$n), drop = TRUE)
  summary_rows <- do.call(rbind, lapply(split_rows, function(d) data.frame(
    scenario = sub("scenario_", "", d$scenario[[1]]),
    engine = d$engine[[1]],
    agents = d$n[[1]],
    replicates = nrow(d),
    median_sim_s = median(d$simulate_seconds),
    q25_sim_s = q(d$simulate_seconds, 0.25),
    q75_sim_s = q(d$simulate_seconds, 0.75)
  )))
  summary_rows <- summary_rows[order(summary_rows$scenario, summary_rows$agents,
                                     summary_rows$median_sim_s), ]
  knitr::kable(summary_rows, digits = 3,
    col.names = c("Scenario", "Engine", "Agents", "Runs", "Median simulation (s)",
                  "Q1 (s)", "Q3 (s)"),
    row.names = FALSE)
}
```

## Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel, and
the scenarios sit side by side for each engine.

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 5.5
if (has_full) {
  full$engine <- factor(full$engine, levels = engine_levels)
  full$agents <- factor(full$n, levels = size_levels,
                        labels = c("10,000 agents", "100,000 agents"))
  print(
    ggplot2::ggplot(full, ggplot2::aes(engine, simulate_seconds, fill = scenario_label)) +
      ggplot2::geom_boxplot(width = 0.7, outlier.alpha = 0.25,
                            position = ggplot2::position_dodge(width = 0.8)) +
      ggplot2::facet_wrap(~ agents, scales = "free_y") +
      ggplot2::scale_y_log10() +
      ggplot2::scale_fill_brewer(palette = "Set2") +
      ggplot2::labs(
        x = NULL, fill = "Scenario",
        y = "Simulation wall time\n(seconds, log scale)",
        subtitle = "Wall time distribution for 100-day simulations"
        ) +
      ggplot2::theme_minimal(base_size = 16) +
      ggplot2::theme(
        legend.position = "bottom",
        axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
        )
  )
}
```

## Cost of added complexity

### Run time

Each later scenario is compared with scenario 00 replicate by replicate: the
engine, population size, seed, network, and calibration are the same, and
only the model differs. Values above one mean that the more complex scenario
took longer.

```{r}
#| label: complexity-time
if (has_full && baseline %in% full$scenario) {
  base <- full[full$scenario == baseline, c("engine", "n", "seed", "simulate_seconds")]
  names(base)[4] <- "baseline_seconds"
  later <- merge(full[full$scenario != baseline, ], base, by = c("engine", "n", "seed"))
  if (nrow(later)) {
    later$ratio <- later$simulate_seconds / later$baseline_seconds
    groups <- split(later, list(later$scenario, later$engine, later$n), drop = TRUE)
    complexity <- do.call(rbind, lapply(groups, function(d) data.frame(
      scenario = sub("scenario_", "", d$scenario[[1]]),
      engine = as.character(d$engine[[1]]),
      agents = d$n[[1]],
      median = median(d$ratio),
      q25 = q(d$ratio, .25),
      q75 = q(d$ratio, .75)
    )))
    complexity <- complexity[order(complexity$scenario, complexity$agents, complexity$median), ]
    knitr::kable(complexity, digits = 2,
      col.names = c("Scenario", "Engine", "Agents", "Median time / scenario 00", "Q1", "Q3"),
      row.names = FALSE)
  }
}
```

Run time does not isolate the cost of the extra machinery. Scenario 01's
vaccine protects about a quarter of the population, which cuts the median
attack rate from about 0.38 to 0.11 at 10,000 agents and from 0.06 to 0.013
at 100,000. Engines whose work tracks the number of active infections (EoN,
epydemic, epiworldR, and ixa) get faster for that reason alone. Covasim's
vectorized daily update touches every agent regardless of the outbreak's
size, so its time barely moves. A ratio below one therefore does not mean
the vaccine is free. It means that the engine's handling of the vaccine costs
less than the smaller outbreak saves. A ratio above one would mean the
feature itself is expensive, as `distribute_tool_to_set()` made it for
epiworldR (see [Design and interpretation](#design-and-interpretation)).

### Code required to define the model

A rough measure of effort: how much code each engine needs to express each
scenario. *Model lines* counts non-blank, non-comment lines that build the
population and network, define the disease and interventions, and run the
simulation. Argument parsing, edge-file reading, timing, and writing results
are excluded because every runner shares them. *Files* counts the files a user
writes, including build manifests. They are counted from the runner sources
when this report is rendered, using the regions listed in each scenario's
`code_regions.yml`.

```{r}
#| label: code-effort
# Each region runs from the first line matching `from` to the next line
# matching `to` (both inclusive). The anchors live in code_regions.yml rather
# than as markers in the runners because runner sources are part of the cache
# fingerprint.
count_model_lines <- function(spec, scenario) {
  path <- file.path(scenario, spec$files[[1]])
  lines <- readLines(path, warn = FALSE)
  keep <- logical(length(lines))
  for (r in spec$regions) {
    to <- if (is.null(r$to)) r$from else r$to
    start <- grep(r$from, lines)[1]
    if (is.na(start)) stop("Anchor not found in ", path, ": ", r$from)
    end <- start - 1L + grep(to, lines[start:length(lines)])[1]
    if (is.na(end)) stop("Anchor not found in ", path, ": ", to)
    keep[start:end] <- TRUE
  }
  code <- trimws(lines[keep])
  timing <- "perf_counter|proc\\.time|Instant::now|elapsed"
  sum(nzchar(code) & !grepl("^(#|//|\"\"\")", code) & !grepl(timing, code))
}

effort <- do.call(rbind, lapply(scenario_levels, function(scenario) {
  specs <- yaml::read_yaml(file.path(scenario, "code_regions.yml"))
  data.frame(
    scenario = scenario,
    engine = names(specs),
    language = vapply(specs, `[[`, "", "language"),
    files = vapply(specs, function(x) length(x$files), 0L),
    model_lines = vapply(specs, count_model_lines, 0L, scenario = scenario)
  )
}))
lines_wide <- reshape(effort[c("engine", "scenario", "model_lines")],
                      idvar = "engine", timevar = "scenario", direction = "wide")
effort_table <- merge(unique(effort[c("engine", "language")]), lines_wide, by = "engine")
line_columns <- paste0("model_lines.", scenario_levels)
later_columns <- setdiff(line_columns, paste0("model_lines.", baseline))
for (column in later_columns) {
  effort_table[[sub("model_lines", "added", column)]] <-
    effort_table[[column]] - effort_table[[paste0("model_lines.", baseline)]]
}
files_by_engine <- tapply(effort$files, effort$engine, max)
effort_table$files <- files_by_engine[effort_table$engine]
effort_table <- effort_table[order(effort_table[[line_columns[[1]]]]),
  c("engine", "language", "files", line_columns, sub("model_lines", "added", later_columns))]
knitr::kable(effort_table,
  col.names = c("Engine", "Language", "Files",
                paste("Lines,", sub("scenario_", "scenario ", scenario_levels)),
                paste("Added in", sub("scenario_", "scenario ", sub("model_lines.", "", later_columns)))),
  row.names = FALSE)
```

Model lines vary with how much of the model an engine provides built in. For
example, EoN describes transitions as rate graphs, Covasim needs its native
parameters overridden to restrict it to SEIRH, and the ixa runner writes its
daily step by hand. The Python engines share one runner file per scenario,
and ixa needs a Cargo project that is compiled before it runs.

For the vaccine in scenario 01, EoN and epydemic only need a compartment with
no transitions, Covasim has a vaccine intervention but it is leaky, so an
all-or-nothing vaccine takes two targeted doses, epiworldR has a tool that
removes susceptibility, and ixa adds an agent property that its hand-written
transmission step has to consult.

## Speed relative to epiworldR

Ratios are matched by scenario, population size, and replicate seed. Values
above one mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
if (has_full && any(full$engine == "epiworldR")) {
  epiworld <- full[full$engine == "epiworldR", c("scenario", "n", "seed", "simulate_seconds")]
  names(epiworld)[4] <- "epiworld_seconds"
  ratios <- merge(full[full$engine != "epiworldR", ], epiworld, by = c("scenario", "n", "seed"))
  ratios$speed_ratio <- ratios$simulate_seconds / ratios$epiworld_seconds
  ratio_groups <- split(ratios, list(ratios$scenario, ratios$engine, ratios$n), drop = TRUE)
  ratio_table <- do.call(rbind, lapply(ratio_groups, function(d) data.frame(
    scenario = sub("scenario_", "", d$scenario[[1]]),
    engine = as.character(d$engine[[1]]),
    agents = d$n[[1]],
    median = median(d$speed_ratio),
    q25 = q(d$speed_ratio, .25),
    q75 = q(d$speed_ratio, .75)
  )))
  ratio_table <- ratio_table[order(ratio_table$scenario, ratio_table$agents, ratio_table$engine), ]
  knitr::kable(ratio_table, digits = 2,
    col.names = c("Scenario", "Engine", "Agents", "Median time / epiworldR", "Q1", "Q3"),
    row.names = FALSE)
}
```

## Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show final attack rate and peak hospitalization load, and, where
a scenario vaccinates, the share of agents the vaccine protected; differences
also expose the engines' non-identical time semantics and Covasim's native
symptom/severe bookkeeping.

```{r}
#| label: outcomes-table
if (has_full) {
  full$attack_rate <- 1 - full$final_susceptible / full$n
  full$protected_share <- full$vaccine_protected / full$n
  outcomes <- do.call(rbind, lapply(
    split(full, list(full$scenario, full$engine, full$n), drop = TRUE),
    function(d) data.frame(
      scenario = d$scenario[[1]],
      scenario_label = d$scenario_label[[1]],
      engine = as.character(d$engine[[1]]),
      n = d$n[[1]],
      attack_rate = median(d$attack_rate),
      peak_hospitalized = median(d$peak_hospitalized),
      protected_share = if (all(is.na(d$protected_share))) NA else median(d$protected_share, na.rm = TRUE)
    )
  ))
  outcomes <- outcomes[order(outcomes$scenario, outcomes$n, outcomes$engine), ]
  shown <- outcomes[c("scenario", "engine", "n", "attack_rate", "peak_hospitalized", "protected_share")]
  shown$scenario <- sub("scenario_", "", shown$scenario)
  knitr::kable(shown, digits = 3,
    col.names = c("Scenario", "Engine", "Agents", "Median final attack rate",
                  "Median peak hospitalized", "Median share protected"),
    row.names = FALSE)
}
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (has_full) {
  outcomes$agents <- factor(outcomes$n, levels = size_levels,
                             labels = c("10,000", "100,000"))
  outcomes$engine <- factor(outcomes$engine, levels = engine_levels)
  print(
    ggplot2::ggplot(outcomes, ggplot2::aes(engine, attack_rate, fill = scenario_label)) +
      ggplot2::geom_col(width = 0.7, position = ggplot2::position_dodge(width = 0.8)) +
      ggplot2::facet_wrap(~ agents, scales = "free_y") +
      ggplot2::scale_y_continuous(limits = c(0, NA)) +
      ggplot2::scale_fill_brewer(palette = "Set2") +
      ggplot2::labs(x = NULL, y = "Median final attack rate", fill = "Scenario",
                    subtitle = "The calibrated outcomes are closely aligned within each scenario and population size") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::theme(legend.position = "bottom")
  )
}
```

## Design and interpretation

The scenario 00 disease graph is susceptible $\rightarrow$ exposed $\rightarrow$
infectious $\rightarrow$ recovered, with a competing infectious $\rightarrow$
hospitalized $\rightarrow$ recovered branch. Mean latent and infectious
periods are 4 and 7 days, lifetime hospitalization probability is 5%, and the
hospital stay is 7 days. Initial infections are 100 in the full profile. They
start infectious in every engine except epiworldR, whose R API cannot seed a
custom model's initial cases in a state other than the one new infections
enter, so they start exposed there.

Scenario 01 adds an all-or-nothing vaccine before the first day: 30% of agents
are vaccinated at random, and each is fully protected with probability 0.8, so
about 24% of the population can never be infected. Every engine expresses the
vaccine in its own idiom, described in
[scenario_01/README.md](scenario_01/README.md). One of these choices changed
the results substantially: epiworldR's `distribute_tool_to_set()` copies the
list of target agents once per agent it reaches, so placing the tool on 24,000
named agents took about 4.6 seconds per run at 100,000 agents. The runner
instead draws the number of protected agents and lets epiworld place the tool
at random, which is statistically the same vaccine. The epydemic runner draws
its vaccinated agents inside the timed simulation call, where epydemic sets
initial compartments; this adds well under a millisecond.

The shared analytic transmission mapping produced different realized attack
rates across frameworks. Calibration therefore selected size-specific factors
for restricted Covasim (0.715 at 10,000 agents and 0.710 at 100,000) and for
ixa (0.983 and 0.988). epiworldR needs none: its uncalibrated median attack
rates match EoN's exact continuous-time results. All 100 replicates in each
adjusted cell were then rerun. These empirical corrections are intentionally
population-specific; they align benchmark outcomes but mean that none of the
calibrated engines has a strictly analytic $R_0$ of 2. Scenario 01 reuses the
same factors, because they correct the engines' transmission semantics rather
than anything specific to a scenario. The values live in each scenario's
`scenario.toml` and participate in the cache fingerprint.

epiworldR, epydemic, and ixa use synchronous daily transitions. The ixa
runner schedules one plan per day on ixa's plan queue, samples transmission
along ixa's built-in contact-network edges, and stores disease status as an
indexed entity property; its competing infectious $\rightarrow$
hospitalized/recovered step reuses epiworldR's roulette rule.

The two fastest engines organize the daily work differently. ixa's step visits
only exposed, infectious, and hospitalized agents, found through its indexed
status property, and tries transmission outward from each infectious agent to
its susceptible neighbours, so its cost follows the size of the outbreak.
epiworldR's queuing system skips agents with no infected contact, but each
queued susceptible agent (every neighbour of an exposed, infectious, or
hospitalized agent) scans all of its neighbours for infectious ones, which is
roughly ten times as many neighbour visits. Its daily loop also checks every
agent's queue flag, about 12% of its run time at 100,000 agents. An exact
push-style alternative for epiworld is proposed in
[UofUEpiBio/epiworld#264](https://github.com/UofUEpiBio/epiworld/issues/264). Both runners
are single-threaded, and all five engines treat the shared edge list as an
undirected graph. EoN uses
continuous-time hazards, simulated exactly with its event-driven
`fast_simple_contagion` algorithm. Covasim retains its native exposed,
infectious, symptomatic, and severe bookkeeping, with severe prevalence used
as the hospitalization proxy. Its runner disables waning immunity and fixes
individual transmissibility and viral load to one, making recovered people
permanently removed and removing those sources of heterogeneity. Covasim does
not expose a public switch to remove the remaining symptom/severity
bookkeeping. Thus this report measures restricted, representative framework
throughput under aligned network and disease targets; it does not claim
bit-for-bit epidemiological equivalence.

The sparse contact network fixes mean degree rather than literal graph density.
At 10,000 and 100,000 agents its densities are approximately 0.001 and 0.0001,
respectively, while every agent still has about ten contacts. This prevents the
edge count, runtime, and memory from growing quadratically.

More details regarding the project setup can be found in [setup.md](./setup.md),
and the steps for adding a scenario in [scenarios.md](./scenarios.md).

## Reproducibility and cache

Every successful replicate is an atomic JSON cache record keyed by scenario,
model configuration, runner source, engine version, and contact-network
SHA-256. Rerunning `make benchmark` schedules only missing or stale records,
and editing one scenario's runners leaves the other scenarios' records valid.
Replicate seeds do not depend on the scenario, which is what makes the
replicate-by-replicate scenario comparison possible. The default
worker count is one, and common native math-library thread counts are pinned to
one. Concurrency is opt-in through the `N_THREADS` environment variable.
Concurrent replicates contend for memory bandwidth and performance cores, and
the penalty differs by engine, so timings are comparable only across runs at
the same concurrency. The platform and worker count for the run behind this
report are shown in the table above; the benchmark runs inside the container
defined in `.devcontainer/` (see [setup.md](./setup.md)), which pins every
engine's toolchain.

```{r}
#| label: versions
if (nrow(all_results)) {
  versions <- unique(all_results[c("engine", "engine_version")])
  knitr::kable(versions[order(versions$engine), ],
               col.names = c("Engine", "Recorded version"), row.names = FALSE)
}
```

## References
