---
title: "Scenario 04: SEIRH on a collapsed GeoPops network"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_04")
full <- bench$full
```

[Back to the project overview](../README.md) ·
[Scenario 00 report](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md)'s model on a real contact network
instead of the shared Watts–Strogatz graph: the model, its parameters, the
100 seed cases, and the runners in [`runners/`](runners) are copies of
scenario 00's. Two engines are added here rather than in scenario 00 because
they need extra build steps and are only worth that cost once: FRED (a
compiled, population-and-network-driven simulator) and Agents.jl (Julia).
This scenario also asks a different question from scenario 00: does the
picture change on a graph with real degree heterogeneity, rather than
Watts–Strogatz's near-constant degree?

## Model

See [scenario 00](../scenario_00/README.md#model) for the disease model, its
parameters, and the timing conventions.

### The network

Every earlier scenario shares a synthetic Watts–Strogatz network (mean
degree 10, rewiring probability 0.05, so nearly every agent has close to the
same number of contacts). This scenario instead uses a network built from
[GeoPops](https://github.com/GeoPopsHub){target="_blank"}'
[Spartanburg County, South Carolina synthetic
population](https://github.com/GeoPopsHub/sc_spartanburg_measles){target="_blank"}:
a real, degree-heterogeneous contact structure, at the cost of not being able
to vary population size independently of the source data.

[`scripts/generate_network.py`](../scripts/generate_network.py) downloads
GeoPops' four contact-layer adjacency matrices (household, workplace, school,
and group-quarters, each a checksum-verified `MatrixMarket` file pinned to a
commit), keeps the induced subgraph on the first 200,000 rows of each, and
unions the four layers into one undirected, unweighted graph: 376,526 edges,
mean degree 3.77. Layer identity, edge weights, demographic attributes, and
contacts touching excluded rows are discarded; only which pairs of the
200,000 people are ever in contact survives. [`scenario.toml`](scenario.toml)
records the exact commit and per-layer checksums.

| Parameter (`scenario.toml`) | Value | Meaning |
|:-----------------------------|------:|:--------|
| Population                   | 200,000 | GeoPops rows kept |
| Edges                        | 376,526 | after union and dedup |
| Mean degree                  | 3.77 | observed, not fixed |
| Replicates                   | 20 | fewer than scenario 00; the slowest engines take seconds per replicate here |

Like scenario 03, this scenario runs one replicate at a time
(`workers = 1` in its `[design]` table), so its timings are comparable to
every other scenario's default sequential runs.

### Calibration

Scenario 00's per-engine transmission multipliers were derived from full
runs at 10,000 and 100,000 agents on the Watts–Strogatz network, and do not
carry over to a different topology at a different size. Every engine
therefore runs uncalibrated here (`transmission_multiplier` defaults to 1.0)
until a full run's median attack rates give real factors to calibrate
against; see [`scenario.toml`](scenario.toml).

### Engine implementations

The nine engines from [scenario 00](../scenario_00/README.md#engine-implementations)
run the same model unchanged, reading this scenario's edge list instead.
Two more are added:

- **FRED** ([PublicHealthDynamicsLab/FRED](https://github.com/PublicHealthDynamicsLab/FRED){target="_blank"}):
  a compiled, DSL-driven simulator with no edge-list network primitive
  comparable to the other engines' graphs. The runner
  ([`run_FRED.py`](runners/run_FRED.py)) builds a synthetic population of
  200,000 single-person households (so FRED's own household mixing adds no
  contacts beyond the network) and loads the benchmark's edges verbatim as
  `Contact.add_edge` properties on FRED's `Network` group type, with
  transmission restricted to that network. Its contact model differs from
  the other engines' constant per-edge hazard: each infectious agent draws a
  number of successful contacts per day proportional to its degree, so the
  runner derives a per-day, per-edge probability that reproduces the same
  target $R_0$ over the mean infectious period
  ([`daily_edge_probability`](runners/run_FRED.py)).
- **Agents.jl** ([JuliaDynamics/Agents.jl](https://github.com/JuliaDynamics/Agents.jl){target="_blank"}):
  a `StandardABM` with a per-agent status and a per-day `model_step!` that
  scans each infectious agent's adjacency list, matching the synchronous
  daily semantics of epiworldR, epydemic, and individual. The adjacency list
  and transmission math are written for this benchmark, since Agents.jl has
  no built-in epidemic model.

**MEmilio is deliberately excluded.** Its ABM has no edge-list network
primitive either — agents interact only through shared Locations (household,
work, school, ...) — so matching this network would need on the order of
376,000 synthetic two-person Locations, of unproven performance, on top of an
8-compartment, viral-load-driven disease model with no simple SEIRH mapping.
See [issue #11](https://github.com/UofUEpiBio/epiworld-benchmark/issues/11).

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 12
#| fig-height: 5
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by replicate seed. Values above one mean that epiworldR
completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

The simulation time above covers what an engine has to redo for every
replicate on the same network (see
[what is measured](../README.md#run-time)). *Build* is the rest of the
setup: turning the edge list into the engine's population and network, which
a user does once per network. *Build + simulate* is the time to a first
result once the edge list is in memory.

```{r}
#| label: phases
table_phases(bench)
```

### The epiworld family

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate and peak hospitalization load.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "Uncalibrated outcomes on the collapsed GeoPops network"))
}
```

### Code required to define the model

A rough measure of effort: how much code each engine needs to express the
model. The counting rules are described in the
[project overview](../README.md#measuring-implementation-effort), and the
counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench)
```

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
