---
title: "Scenario 04: SEIRH on a GeoPops contact network"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_04")
```

[Back to the project overview](../README.md) ·
[Methods](../docs/methods.md) · [Results](../docs/results.md) ·
[Scenario 00](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md)'s model on a real contact network
instead of the shared Watts–Strogatz graph. The model, its parameters, the
100 seed cases, and the runners in [`runners/`](runners) are scenario 00's.
It asks whether the picture changes on a graph with real degree
heterogeneity rather than Watts–Strogatz's near-constant degree.

## Model

### The network

The network comes from
[GeoPops](https://github.com/GeoPopsHub){target="_blank"}'
[Spartanburg County, South Carolina synthetic
population](https://github.com/GeoPopsHub/sc_spartanburg_measles){target="_blank"}:
a real, degree-heterogeneous contact structure, at the cost of not being able
to vary population size independently of the source data.
[`scripts/generate_network.py`](../scripts/generate_network.py) downloads
GeoPops' four contact-layer adjacency matrices (household, workplace, school,
and group quarters, each a checksum-verified `MatrixMarket` file pinned to a
commit), keeps the induced subgraph on the first 200,000 rows of each, and
unions the four layers into one undirected, unweighted graph. Of those
200,000 people, 34,135 (17%) have no contacts among the others; they could
never be infected, so they are dropped and the remaining 165,865 renumbered
in row order. Layer identity, edge weights, demographic attributes, and
contacts touching excluded rows are discarded; only which pairs of people
are ever in contact survives. [`scenario.toml`](scenario.toml) records the
exact commit and per-layer checksums. This is a collapsed contact graph, not
a reproduction of GeoPops' multilayer, demographic model.

| Property | Value |
|:---------|------:|
| Agents | 165,865 |
| Edges, after union and deduplication | 376,526 |
| Mean degree (observed) | 4.54 |
| Degree range (median) | 1 to 25 (3) |
| Components (largest) | 17,128 (127,291 agents, 77%) |
| Replicates per engine | 20 |

The largest component caps the attack rate. Because households form cliques
and degrees vary, the shared transmission mapping, which divides $R_0$ by the
mean degree minus one, does not give an $R_0$ of 2 here: the mean excess
degree is 6.9, not 3.5. The mapping is the same for every engine, so the
comparison stays aligned, but the outbreaks are larger than the nominal
$R_0$ suggests. Like scenario 03, this scenario runs one replicate at a time
(`workers = 1` in its `[design]` table).

```{r}
#| label: parameters
table_parameters(bench)
```

## Engine notes

None: every engine runs scenario 00's runner on this edge list.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 7
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by replicate seed. Values above one mean that epiworldR
completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

[Timing](../docs/methods.md#timing) defines build and simulate.

```{r}
#| label: phases
table_phases(bench)
```

### Memory

Median [Q1, Q3] in MiB ([Memory](../docs/methods.md#memory)).

```{r}
#| label: memory
table_memory(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

### Model lines

The runners are scenario 00's, so are their model lines (see
[scenario 00](../scenario_00/README.md#model-lines)).

## Notes

**Calibration.** Every engine runs uncalibrated here (a transmission
multiplier of 1): scenario 00's factors were derived on the Watts–Strogatz
network at other sizes and do not carry over to this topology. Most engines'
medians agree anyway; Covasim's sits highest (see the
[results](../docs/results.md#epidemiological-agreement)). A full run's
medians now exist to calibrate against, if aligned factors are wanted.
