---
title: "Scenario 02: vaccine plus epidemiological outputs"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_02")
```

[Back to the project overview](../README.md) ·
[Methods](../docs/methods.md) · [Results](../docs/results.md) ·
[Scenario 01](../scenario_01/README.md)

[Scenario 01](../scenario_01/README.md), unchanged, plus four outputs that
every engine has to produce from each run, so the comparison includes for
every engine the bookkeeping epiworld always does. The
[results](../docs/results.md#cost-of-added-features) compare each replicate
with scenario 01's, which runs the same epidemic for every engine but
individual: its runner picks each case's source with R's global random
numbers, so only its distribution of outcomes matches.

## Model

| Output | Definition |
|:-------|:-----------|
| Transmission tree | The day, source, and target of every infection. Seed cases have no source. |
| Daily incidence | New infections (S → E) on each day from 1 to 100. |
| Reproductive number | epiworld's definition: each case's number of secondary infections, averaged over the cases infected on each day (a *case* reproductive number, so it falls toward zero near day 100). Day 0 holds the seed cases. |
| Transition matrix | The number of agents moving between each pair of S, E, I, H, and R on each day; day 0 is left out. |

Runners hold the outputs in memory, with the engine's own tools where it
has them. Recording during the run is part of `simulate_seconds`; reading the
outputs out afterwards is `extract_seconds`. Records add `extract_seconds`,
`transmissions`, the transition matrix summed over days 1 to 100
(`transitions_se`, ...), and the daily series (medians in
`results/daily.csv`).

## Engine notes

- **The epiworld family**: nothing to add; the database already holds all
  four outputs (`get_transmissions()`, `get_hist_transition_matrix()`, and
  the reproductive-number getters). The C++ and Python runners aggregate the
  raw vectors themselves.
- **Covasim** reads its infection log; **Starsim** adds `ss.infection_log`.
  Both runners count each agent's dated transitions within the run.
- **EoN**: `return_full_data=True` records each node's status history, from
  which the runner builds the matrix, binning events into days.
- **epydemic** and **Agents.jl**: the model's event handlers or daily step
  append each transmission and transition to lists.
- **ixa**: a data plugin, written by the daily step and by a
  `PropertyChangeEvent` subscription.
- **individual**: the infection process picks each source among the infectious
  neighbours; a second process compares the states on consecutive days.
- **EpiModel**: its transmission matrix (`save.transmat`, filled with
  `set_transmat()` by the infection module) and its `infTime` attribute for
  the tree, and an epidemic statistic per transition, set by the modules.
- **FRED**: its health records log every exposure and state change during
  the run; the runner parses them afterwards.

The runners other than the epiworld family compute the reproductive number
from the tree with epiworld's definition, in a small helper counted as model
code.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

[Timing](../docs/methods.md#timing) defines build and simulate.

```{r}
#| label: phases
table_phases(bench)
```

### Extracting the outputs

Time to read the four outputs out of each engine after the run, not part of
the simulation time.

```{r}
#| label: extraction
table_extraction(bench)
```

### Memory

Median [Q1, Q3] in MiB ([Memory](../docs/methods.md#memory)).

```{r}
#| label: memory
table_memory(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

Share of runs whose outputs agree with their own final counts (every agent
who left S is a seed or a transmission target, every transmission is an
S → E transition, every recovered agent came from I or H):

```{r}
#| label: output-checks
table_output_checks(bench)
```

Daily series, as medians across replicates. Engines that seed their initial
cases exposed (the epiworld family, Covasim, and Starsim) start their
incidence a few days later.

```{r}
#| label: daily-plot
#| fig-width: 10
#| fig-height: 6
if (bench$has_full) print(plot_daily(bench))
```

### Model lines

Counted as described in
[Implementation size](../docs/methods.md#implementation-size); the last
column is what the outputs added to scenario 01. Code that only summarizes
the outputs for the result record is not counted.

```{r}
#| label: code-effort
table_code_effort(bench, baseline = "scenario_01")
```

## Notes

**Extraction is small for every engine but epiworldR and FRED.** FRED's is
parsing its health records. Almost all of epiworldR's is the documented
`plot(get_reproductive_number(model), plot = FALSE)`, which computes a mean,
a standard deviation, and two quantiles for every day and builds a data
frame for each: about 25 ms for 100 days however small the outbreak, against
0.4 ms for a plain `tapply()` on the same data and about 1 ms for
`get_reproductive_number()` itself. The runner keeps the documented call, as
an epiworldR user would write it; a vectorized summary in epiworldR would
remove nearly all of the package's extraction time.
