---
title: "Scenario 02: SEIRH + vaccination + epidemiological outputs"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_02")
baseline <- "scenario_01"
full <- bench$full
# Median paired time ratio against scenario 01, and median extraction time,
# for the inline numbers in the interpretation.
paired <- if (nrow(full)) merge(
  full, setNames(bench$every[bench$every$scenario == baseline, c("engine", "n", "seed", "simulate_seconds")],
                 c("engine", "n", "seed", "baseline_seconds")),
  by = c("engine", "n", "seed")
) else data.frame()
ratio <- function(engine, n) {
  d <- paired[paired$engine == engine & paired$n == n, ]
  if (nrow(d)) sprintf("%.2f", median(d$simulate_seconds / d$baseline_seconds)) else "?"
}
extract_ms <- function(engine, n) {
  d <- full[full$engine == engine & full$n == n, ]
  if (nrow(d)) sprintf("%.1f", 1000 * median(d$extract_seconds)) else "?"
}
```

[Back to the project overview](../README.md) ·
[Scenario 01 report](../scenario_01/README.md)

[Scenario 01](../scenario_01/README.md), unchanged, plus four outputs that
every engine has to produce from each run: the transmission tree, daily
incidence, the reproductive number, and the daily transition matrix.

epiworld records all four during every run, in every scenario, whether or not
anyone reads them. The other engines record less by default, so in scenarios
00 and 01 the epiworld family does bookkeeping that the others skip. Here every
engine has to deliver the same information, so the comparison includes that
bookkeeping for all of them.

The model, network, seeds, run length, and calibration are those of
scenario 01. For every engine but individual, the two scenarios run the same
epidemic for a given size and replicate, and this report compares them
replicate by replicate. individual's runner picks each new case's source
with R's global random numbers, which its disease processes also use, so
from the first such draw its epidemic differs from scenario 01's for the
same seed. The two are still the same model, with the same distribution of
outcomes; only the pairing by seed is lost.

## Model

### Outputs

Each runner produces the following from each run and holds them in memory in
the engine's own data structures. Nothing is written to disk.

| Output | Definition |
|:-------|:-----------|
| Transmission tree | The day, source, and target of every infection. Seed cases have no source. |
| Daily incidence | New infections (S → E) on each day from 1 to 100. |
| Reproductive number | epiworld's definition: each case's number of secondary infections, averaged over the cases infected on each day. Day 0 holds the seed cases. |
| Transition matrix | The number of agents moving between each pair of S, E, I, H, and R on each day. |

- The reproductive number is a *case* reproductive number: it is indexed by
  the day the case was infected, not the day it transmitted. Cases infected
  near day 100 have had little time to transmit, so the series falls toward
  zero at the end of the run in every engine.
- Where an engine's own tools already give an output, the runner uses them.
  Where they do not, the runner builds it from what the engine does offer, the
  way a user of that engine would.
- Transitions on day 0 put the seed cases in place and are left out of the
  daily series and the totals below.

### Timing

`simulate_seconds` covers the simulation call only, as in scenarios 00 and 01.
Any bookkeeping an engine does while it simulates, such as recording
transmissions or transitions, is part of that call and is timed. Reading the
outputs out of the engine afterwards is not: it is reported on its own as
`extract_seconds` and left out of `simulate_seconds` and `setup_seconds`.
As in scenario 01, distributing the vaccine is timed as simulation in every
runner.

This puts the line at the end of the run. An engine that records outputs as
it goes pays for them in the timed call; an engine that reconstructs them
afterwards (Covasim's transition matrix, EoN's matrix from its node
histories) does that part untimed. The extraction table below shows how much
that is.

### Result fields

Runners add these fields to each result record:

- `extract_seconds`: the time taken to extract the outputs after the run.
- `transmissions`: the number of infections with a source, which is the size
  of the transmission tree without the seed cases.
- `transitions_se`, `transitions_ei`, `transitions_ih`, `transitions_ir`,
  `transitions_hr`: the transition matrix summed over days 1 to 100.
- `daily_incidence` (days 1 to 100) and `reproductive_number` (days 0 to
  100, `null` for a day with no cases): the daily series. They stay in the
  cache; `results/daily.csv` holds their medians across replicates.

The five final compartments still sum to `n`, and protected agents who are
never infected are still counted as susceptible.

### Engine implementations

- **epiworldR**: nothing to add to the model. After the run,
  `get_transmissions()` returns the tree and `get_hist_transition_matrix()`
  the matrix. `plot_incidence(plot = FALSE)` and
  `plot(get_reproductive_number(), plot = FALSE)` return the daily series.
  Four lines.
- **epiworld** (C++): the same database, through `get_transmissions()`,
  `get_hist_transition_matrix()`, and `get_reproductive_number()`. The C++ API
  returns raw vectors and a map with one entry per case, so the runner sums
  the matrix's S → E entries and averages the reproductive number by day
  itself.
- **epiworldpy**: the same getters, returning NumPy arrays and dictionaries.
  The runner aggregates them with NumPy.
- **Covasim**: Covasim always keeps an infection log with the source, target,
  and date of every infection, and reports daily new infections. The runner
  reads the log directly: `sim.make_transtree()` wraps the same log but also
  builds a per-case table, and it drops the cases infected by agent 0. Covasim
  has no transition matrix, but it dates each agent's transitions
  (`date_exposed`, `date_infectious`, `date_severe`, `date_recovered`, some of
  them in the future), and the runner counts those that fall within the run.
- **Starsim**: adding its `ss.infection_log` analyzer makes the disease log
  the source and target of every infection, in a NetworkX graph, as it
  happens; daily new infections are a standard result. Like Covasim, Starsim
  has no transition matrix but keeps each agent's transition times, some of
  them scheduled in the future, and the runner counts those that fall within
  the run.
- **EoN**: `return_full_data=True` makes `fast_simple_contagion` record each
  node's status history and every transmission, and return them in a
  `Simulation_Investigation`. The runner builds the matrix from the node
  histories. EoN runs in continuous time, so events are binned into days:
  day *d* covers the interval (*d* − 1, *d*].
- **epydemic**: no built-in record of individual events. The model's event
  handlers append each transmission and each transition to lists, through a
  small `move()` helper that records a transition before changing the
  compartment.
- **ixa**: a data plugin holds the outputs. The daily step, which already
  knows the infectious contact behind each exposure, records the transmission
  when it applies it. A subscription to ixa's `PropertyChangeEvent` for the
  disease status counts every transition. After the run, the runner derives
  incidence and the reproductive number from the plugin.
- **individual**: no built-in record of events or sources. The infection
  process picks each new case's source uniformly among its infectious
  neighbours, as epiworld does, and records it with the day. A second
  process compares the agents in each state at the start of consecutive days
  to count the transitions. After the run, the runner derives the
  reproductive number from the tree.
- **FRED**: its health records, switched on for the disease condition in
  every run (`enable_health_records`, `health_records_run = -1`), log every
  exposure with its source and day and every state change, as FRED
  simulates; writing them is part of the timed run. After the run, the
  runner parses the log into the tree and the transition matrix. Daily
  incidence is the new-exposure column of FRED's own daily report.
- **Agents.jl**: no built-in record of events. The model's properties hold
  two vectors, and the daily step, which already knows the infectious contact
  behind each exposure, appends each transmission and each transition as it
  applies them. After the run, the runner derives the matrix, incidence, and
  reproductive number from them, in a function the warm-up model compiles
  before any timer starts.

The EoN, epydemic, Covasim, Starsim, ixa, individual, FRED, and Agents.jl
runners use epiworld's definition of the reproductive number. Each Python
runner has its own copy of a small helper that computes it from the tree, and
that helper counts as model code.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside the simulation call, including
any recording the engine does during it. The logarithmic scale keeps fast and
slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR finished faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

The simulation time above covers what an engine has to redo for every
replicate on the same network (see
[what is measured](../README.md#run-time)). *Build* is the rest of the
setup: turning the edge list into the engine's population and network, which
a user does once per network. *Build + simulate* is the time to a first
result once the edge list is in memory. Reading the edge file is shown but
left out of it, because its cost depends on each language's file parsing
rather than on the engine.

```{r}
#| label: phases
table_phases(bench)
```

### Extracting the outputs

The time taken to read the four outputs out of each engine after the run.
It is not part of the simulation times above.

```{r}
#| label: extraction
table_extraction(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

The outputs have to agree with each engine's own final counts. Every agent who
left S is either a seed case or the target of a transmission; every
transmission is an S → E transition; and every recovered agent came from I or
H. The table shows the share of runs in which these hold.

```{r}
#| label: output-checks
table_output_checks(bench)
```

The daily series, as medians across replicates. The epiworld family seeds its
initial cases as exposed, and so do Covasim and Starsim, so their incidence
starts a few days later than that of the engines that seed them as infected.

```{r}
#| label: daily-plot
#| fig-width: 10
#| fig-height: 7
daily_plot <- plot_daily(bench)
if (!is.null(daily_plot)) print(daily_plot)
```

## Cost of the outputs

### Run time compared with scenario 01

Each replicate is paired with the scenario 01 replicate that has the same
engine, population size, and seed. Both scenarios run the same epidemic
(for individual, one from the same distribution), so a ratio above one is
the cost of recording the outputs during the run.

```{r}
#| label: time-versus-baseline
table_time_versus(bench, baseline)
```

### Code required to define the model

How much code each engine needs for this scenario, and how much the outputs
added to scenario 01. The counting rules are described in the
[project overview](../README.md#measuring-implementation-effort), and the
counted regions are listed in [`code_regions.yml`](code_regions.yml). Code that
only summarizes the outputs for the result record is not counted.

```{r}
#| label: code-effort
table_code_effort(bench, baseline)
```

## Interpretation

### What recording the outputs costs

For most engines, very little. The epiworld family already records all four
outputs in scenario 01, so its simulation does not change: the C++ runner
takes `r ratio("epiworld", 100000)` times as long as in scenario 01 at 100,000
agents, epiworldR `r ratio("epiworldR", 100000)` times, and epiworldpy
`r ratio("epiworldpy", 100000)` times. Covasim keeps its infection log in every
run too (`r ratio("covasim", 100000)` times). The engines that had to add
recording mostly absorbed it as well. ixa's event subscription and
transmission log leave it at `r ratio("ixa", 100000)` times its scenario 01
time, and epydemic's handlers at `r ratio("epydemic", 100000)` times, where the
list appends are small next to its Python event loop. Agents.jl's
vectors of transmissions and transitions leave it at
`r ratio("Agents.jl", 100000)` times. FRED, whose health records write a
line to a file, and flush it, for every exposure and state change, takes
`r ratio("FRED", 100000)` times as long.

The main exception is **EoN**, which takes `r ratio("EoN", 10000)` times as long at
10,000 agents and `r ratio("EoN", 100000)` times at 100,000. With
`return_full_data=True`, `fast_simple_contagion` appends to every node's
status history, in Python, at every event. **individual** also pays, taking
`r ratio("individual", 10000)` times as long at 10,000 agents and
`r ratio("individual", 100000)` times at 100,000. It has no record of
transitions, so its runner compares the agents in each state at the start of
consecutive days, which means working through full-population bitsets
every day.

Extraction after the run is small for every engine but epiworldR and FRED:
at 100,000 agents it takes `r extract_ms("epiworldR", 100000)` ms for
epiworldR and `r extract_ms("FRED", 100000)` ms for FRED, against
`r extract_ms("EoN", 100000)` ms for EoN, `r extract_ms("Agents.jl", 100000)`
ms for Agents.jl, and under a millisecond for the others. Almost all of
epiworldR's is one call (see below); FRED's is parsing its health records,
which also log the day-0 move of every agent into its initial state.

### Is bookkeeping what separates epiworld and ixa?

No. What epiworld records per event is small: each infection appends one row
to the transmission log, and the daily database update costs time in
proportion to the number of states and tools, not agents. ixa adds the same
bookkeeping here at little cost, and the two engines compare as they do in
scenario 01. What separates them has other causes, the same in every
scenario: ixa rebuilds its context for every replicate, while epiworld looks
up the transmission probability by name for every contact it tries,
re-initializes the whole population at the start of every run, and, with the
vaccine, clones the tool for every agent it protects.
[analysis.md](../analysis.md) measures each.

### An implementation choice that mattered: epiworldR's reproductive-number summary

The documented way to get the daily reproductive number in epiworldR is
`plot(get_reproductive_number(model), plot = FALSE)`. The `plot()` method
computes a mean, a standard deviation, and 2.5% and 97.5% quantiles for every
day, and builds a data frame for each. That takes about 25 ms for 100 days, no
matter how small the outbreak, against 0.4 ms for a plain
`tapply(rt, source_exposure_date, mean)` on the same data frame.
`get_reproductive_number()` itself takes about 1 ms. The runner keeps the
documented call, because that is what an epiworldR user would write. A
vectorized summary in epiworldR would remove nearly all of the package's
extraction time. None of it counts toward the simulation times.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
