---
title: "Scenario 00: SEIRH baseline"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_00")
```

[Back to the project overview](../README.md)

The reference model for the benchmark. Every later scenario adds complexity
to it, and each later scenario's report compares its run time and model code
against this one.

## Model

Susceptible $\rightarrow$ exposed $\rightarrow$ infectious $\rightarrow$
recovered, with a competing infectious $\rightarrow$ hospitalized
$\rightarrow$ recovered branch. Transmission happens along the shared
Watts–Strogatz contact network (mean degree 10, rewiring probability 0.05).
There are no interventions.

| Parameter (`scenario.toml`)   | Value | Meaning                                          |
|:------------------------------|------:|:-------------------------------------------------|
| `target_r0`                   |   2.0 | Early-outbreak $R_0$ used by the analytic mapping |
| `initial_infected`            |   100 | Seed cases (10 in the smoke profile)             |
| `latent_days`                 |   4.0 | Mean exposed period                              |
| `infectious_days`             |   7.0 | Mean infectious period                           |
| `hospitalization_probability` |  0.05 | Lifetime probability that an infectious agent is hospitalized |
| `hospital_days`               |   7.0 | Mean hospital stay                               |

Initial cases start infectious in every engine except epiworldR. Its R API
cannot seed a custom model's initial cases in any state other than the one
new infections enter, so they start exposed there.

### Engine implementations

- **epiworldR**: a custom model built from `add_state()` and the package's
  update-function factories. New infections enter Exposed.
- **Covasim**: native disease progression restricted to SEIRH. Waning is off,
  and individual transmissibility and viral load are fixed. The severe state
  serves as the hospitalization proxy.
- **EoN**: a continuous-time rate graph run with its event-driven
  `fast_simple_contagion` algorithm.
- **epydemic**: a `CompartmentedModel` under `SynchronousDynamics`.
- **ixa**: one plan per day, ixa's built-in contact network, and an indexed
  disease-status property. The competing infectious transitions reuse
  epiworldR's roulette rule.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate and peak hospitalization load.
Differences also expose the engines' non-identical time semantics and
Covasim's native symptom/severe bookkeeping.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The calibrated outcomes are closely aligned within each population size"))
}
```

### Code required to define the model

A rough measure of effort: how much code each engine needs to express the
model. The counting rules are described in the
[project overview](../README.md#measuring-implementation-effort), and the
counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench)
```

Model lines vary with how much of the model an engine provides built in. For
example, EoN describes transitions as rate graphs, Covasim needs its native
parameters overridden to restrict it to SEIRH, and the ixa runner writes its
daily step by hand. The Python engines share one runner file, and ixa needs a
Cargo project that is compiled before it runs.

## Interpretation

### Calibration

The shared analytic transmission mapping produced different realized attack
rates across frameworks. Calibration therefore selected size-specific factors
for restricted Covasim (0.715 at 10,000 agents and 0.710 at 100,000) and for
ixa (0.983 and 0.988). epiworldR needs none: its uncalibrated median attack
rates match EoN's exact continuous-time results. All 100 replicates in each
adjusted cell were then rerun. These empirical corrections are intentionally
population-specific; they align benchmark outcomes but mean that none of the
calibrated engines has a strictly analytic $R_0$ of 2. The values live in
[`scenario.toml`](scenario.toml) and participate in the cache fingerprint.

Covasim's factors were calibrated after restricting it to fixed transmission
and permanent immunity. ixa shares the synchronous daily semantics of
epiworldR and epydemic, but its uncalibrated attack rates sat above the other
engines: it seeds the initial cases as infectious and samples each infectious
contact independently rather than with epiworld's roulette.

### Why the two fastest engines differ

ixa's step visits only exposed, infectious, and hospitalized agents, found
through its indexed status property, and tries transmission outward from each
infectious agent to its susceptible neighbours, so its cost follows the size
of the outbreak. epiworldR's queuing system skips agents with no infected
contact, but each queued susceptible agent (every neighbour of an exposed,
infectious, or hospitalized agent) scans all of its neighbours for infectious
ones, which is roughly ten times as many neighbour visits. Its daily loop also
checks every agent's queue flag, about 12% of its run time at 100,000 agents.
An exact push-style alternative for epiworld is proposed in
[UofUEpiBio/epiworld#264](https://github.com/UofUEpiBio/epiworld/issues/264).

### Equivalence across engines

epiworldR, epydemic, and ixa use synchronous daily transitions, while EoN
uses continuous-time hazards, simulated exactly. Covasim retains its native
exposed, infectious, symptomatic, and severe bookkeeping, with severe
prevalence used as the hospitalization proxy. Its runner disables waning
immunity and fixes individual transmissibility and viral load to one, making
recovered people permanently removed and removing those sources of
heterogeneity. Covasim does not expose a public switch to remove the
remaining symptom/severity bookkeeping. Thus this report measures restricted,
representative framework throughput under aligned network and disease
targets; it does not claim bit-for-bit epidemiological equivalence.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
