---
title: "Scenario 00: SEIRH baseline"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_00")
```

[Back to the project overview](../README.md)

The reference model for the benchmark. Every later scenario adds complexity
to it, and each later scenario's report compares its run time and model code
against this one.

## Model

Susceptible $\rightarrow$ exposed $\rightarrow$ infectious $\rightarrow$
recovered, with a competing infectious $\rightarrow$ hospitalized
$\rightarrow$ recovered branch. Transmission happens along the shared
Watts–Strogatz contact network (mean degree 10, rewiring probability 0.05).
There are no interventions.

| Parameter (`scenario.toml`)   | Value | Meaning                                          |
|:------------------------------|------:|:-------------------------------------------------|
| `target_r0`                   |   2.0 | Early-outbreak $R_0$ used by the analytic mapping |
| `initial_infected`            |   100 | Seed cases (10 in the smoke profile)             |
| `latent_days`                 |   4.0 | Mean exposed period                              |
| `infectious_days`             |   7.0 | Mean infectious period                           |
| `hospitalization_probability` |  0.05 | Lifetime probability that an infectious agent is hospitalized |
| `hospital_days`               |   7.0 | Mean hospital stay                               |

Initial cases start infectious in every engine except the epiworld family.
epiworldR's API cannot seed a custom model's initial cases in any state other
than the one new infections enter, so they start exposed there. The C++ and
Python runners do the same, so that all three epiworld runners build exactly
the same model.

### Engine implementations

- **epiworld** (C++): a custom model built from `add_state()` and the
  library's update-function factories, `sampler::make_update_susceptible()` and
  `new_state_update_transition()`. New infections enter Exposed. It is
  compiled with the flags R uses for epiworldR (`-O2 -DNDEBUG`, double
  precision).
- **epiworldR**: the same model through the R wrapper, with
  `update_fun_susceptible()` and `update_fun_rate()`.
- **epiworldpy**: the same model through the Python wrapper, with
  `UpdateFun.susceptible()` and `UpdateFun.rate()`. These were added for this
  benchmark in
  [UofUEpiBio/epiworldpy#17](https://github.com/UofUEpiBio/epiworldpy/pull/17).
  Before, a custom model's other states needed Python callbacks, which
  epiworld calls for every agent every day. epiworldpy compiles epiworld with
  its default single-precision `float`, and with the same seed it still gives
  the same epidemics as the other two.
- **Covasim**: native disease progression restricted to SEIRH. Waning is off,
  and individual transmissibility and viral load are fixed. The severe state
  serves as the hospitalization proxy.
- **EoN**: a continuous-time rate graph run with its event-driven
  `fast_simple_contagion` algorithm.
- **epydemic**: a `CompartmentedModel` under `SynchronousDynamics`.
- **ixa**: one plan per day, ixa's built-in contact network, and an indexed
  disease-status property. The competing infectious transitions reuse
  epiworldR's roulette rule.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 12
#| fig-height: 5
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### The epiworld family

The three epiworld runners build the same model on the same C++ core, and
in this scenario they produce identical epidemics for every seed. This table
compares the two wrappers with the C++ runner, replicate by replicate.

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate and peak hospitalization load.
Differences also expose the engines' non-identical time semantics and
Covasim's native symptom/severe bookkeeping.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The calibrated outcomes are closely aligned within each population size"))
}
```

### Code required to define the model

A rough measure of effort: how much code each engine needs to express the
model. The counting rules are described in the
[project overview](../README.md#measuring-implementation-effort), and the
counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench)
```

Model lines vary with how much of the model an engine provides built in. For
example, EoN describes transitions as rate graphs, Covasim needs its native
parameters overridden to restrict it to SEIRH, and the ixa runner writes its
daily step by hand. The Python engines share one runner file, and ixa needs a
Cargo project that is compiled before it runs.

## Interpretation

### Calibration

The shared analytic transmission mapping produced different realized attack
rates across frameworks. Calibration therefore selected size-specific factors
for restricted Covasim (0.715 at 10,000 agents and 0.710 at 100,000) and for
ixa (0.983 and 0.988). epiworldR needs none: its uncalibrated median attack
rates match EoN's exact continuous-time results, and so do epiworld's and
epiworldpy's, which run the same model. All 100 replicates in each
adjusted cell were then rerun. These empirical corrections are intentionally
population-specific; they align benchmark outcomes but mean that none of the
calibrated engines has a strictly analytic $R_0$ of 2. The values live in
[`scenario.toml`](scenario.toml) and participate in the cache fingerprint.

Covasim's factors were calibrated after restricting it to fixed transmission
and permanent immunity. ixa shares the synchronous daily semantics of
epiworldR and epydemic, but its uncalibrated attack rates sat above the other
engines: it seeds the initial cases as infectious and samples each infectious
contact independently rather than with epiworld's roulette.

### The language layer costs nothing measurable in the simulation

Neither wrapper is slower than the C++ runner. At 10,000 agents epiworldR
and epiworldpy are even 7-9% faster, and at 100,000 agents they take about
0.015 seconds against the C++ runner's 0.026, although all three run the
same compiled code on the same model. The C++ runner is not doing more work: run twice in the same
process, its second run takes 0.016 seconds. Raising glibc's `mmap`
threshold (`MALLOC_MMAP_THRESHOLD_`), so that large blocks come from the
ordinary heap, also brings its first run down to 0.017 seconds. The
difference is where the model's large per-agent arrays land in memory. In
the C++ runner, each array gets its own fresh, page-aligned `mmap` region.
The R and Python processes have already grown and freed a large heap while
starting up and reading the edge list, so the arrays land there instead.
The run makes no page faults in either case, which points to cache or TLB
effects of the placement rather than to first-touch cost. The runner is
timed as a C++ user would write it, so the published numbers keep this
effect. Allocating those arrays together would likely remove it upstream.

Outside the simulation call the ordering flips. The C++ runner reads the
edge list and builds the model fastest, so its `total_seconds` (which
excludes interpreter startup) is the lowest of the three.

### Why the two fastest engines differ

epiworld 0.16.1 made network transmission push-based when that is cheaper:
infectious agents push infection to their neighbours instead of every queued
susceptible agent scanning all of its neighbours for infectious ones (the
change proposed in
[UofUEpiBio/epiworld#264](https://github.com/UofUEpiBio/epiworld/issues/264)).
Together with the other changes through 0.17.0, this made epiworldR about
four times faster here than 0.15.1.0 was at 100,000 agents (median 0.062 to
0.016 seconds). In push mode, both engines' daily cost now follows the
outbreak rather than the population: each visits only the exposed,
infectious, and hospitalized agents, and tries transmission outward from the
infectious ones.
ixa is still about twice as fast at 100,000 agents. This report does not
profile where the remaining gap comes from.

### Equivalence across engines

epiworldR, epydemic, and ixa use synchronous daily transitions, while EoN
uses continuous-time hazards, simulated exactly. Covasim retains its native
exposed, infectious, symptomatic, and severe bookkeeping, with severe
prevalence used as the hospitalization proxy. Its runner disables waning
immunity and fixes individual transmissibility and viral load to one, making
recovered people permanently removed and removing those sources of
heterogeneity. Covasim does not expose a public switch to remove the
remaining symptom/severity bookkeeping. Thus this report measures restricted,
representative framework throughput under aligned network and disease
targets; it does not claim bit-for-bit epidemiological equivalence.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
