---
title: "Scenario 00: SEIRH baseline"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_00")
full <- bench$full
median_of <- function(engine, n, column = "simulate_seconds") {
  d <- full[full$engine == engine & full$n == n, ]
  if (!nrow(d)) return(NA)
  if (column == "build_seconds") median(d$setup_seconds - d$read_seconds) else median(d[[column]])
}
s3 <- function(x) if (is.na(x)) "?" else sprintf("%.3f", x)
ratio <- function(a, b, n) sprintf("%.1f", median_of(a, n) / median_of(b, n))
```

[Back to the project overview](../README.md)

The reference model for the benchmark. Every later scenario adds complexity
to it, and each later scenario's report compares its run time and model code
against this one.

## Model

Susceptible $\rightarrow$ exposed $\rightarrow$ infectious $\rightarrow$
recovered, with a competing infectious $\rightarrow$ hospitalized
$\rightarrow$ recovered branch. Transmission happens along the shared
Watts–Strogatz contact network (mean degree 10, rewiring probability 0.05).
There are no interventions.

| Parameter (`scenario.toml`)   | Value | Meaning                                          |
|:------------------------------|------:|:-------------------------------------------------|
| `target_r0`                   |   2.0 | Early-outbreak $R_0$ used by the analytic mapping |
| `initial_infected`            |   100 | Seed cases (10 in the smoke profile)             |
| `latent_days`                 |   4.0 | Mean exposed period                              |
| `infectious_days`             |   7.0 | Mean infectious period                           |
| `hospitalization_probability` |  0.05 | Lifetime probability that an infectious agent is hospitalized |
| `hospital_days`               |   7.0 | Mean hospital stay                               |

Initial cases start infectious in every engine except the epiworld family,
Starsim, and FRED. epiworldR's API cannot seed a custom model's initial cases
in any state other than the one new infections enter, so they start exposed
there. The C++ and Python runners do the same, so that all three epiworld
runners build exactly the same model. Starsim's SEIR puts its seed cases where
new infections go, which is also exposed, and so does FRED's import.

The simulation time covers everything an engine has to redo for another
replicate on the same network (see
[what is measured](../README.md#run-time)). Seeding the initial cases is
part of it in every runner, because the seeds differ in every run and
epiworld places them inside `run()`, which also re-initializes the whole
population. EoN's timer covers building its initial status dictionary,
individual's covers building its state variable, and the other engines seed
inside their simulation call. ixa's `execute()` runs a context only once, so
its timer also covers building the context: the population, the network, and
the index on disease status; it then seeds in a plan at time 0. Starsim's
`Sim` also runs only once, so its timer covers building the disease, network,
and `Sim` objects and `sim.init()`, which builds the agent and network arrays
and seeds the initial cases. FRED runs one replicate per process and rebuilds
everything each time; its timer covers building its places, population, and
network as well as the days (see its engine description below).

[Scenario 03](../scenario_03/README.md) runs this model at 1,000,000
agents.

### Engine implementations

- **epiworld** (C++): a custom model built from `add_state()` and the
  library's update-function factories, `sampler::make_update_susceptible()` and
  `new_state_update_transition()`. New infections enter Exposed. It is
  compiled with the flags R uses for epiworldR (`-O2 -DNDEBUG`, double
  precision).
- **epiworldR**: the same model through the R wrapper, with
  `update_fun_susceptible()` and `update_fun_rate()`.
- **epiworldpy**: the same model through the Python wrapper, with
  `UpdateFun.susceptible()` and `UpdateFun.rate()`. These were added for this
  benchmark in
  [UofUEpiBio/epiworldpy#17](https://github.com/UofUEpiBio/epiworldpy/pull/17).
  Before, a custom model's other states needed Python callbacks, which
  epiworld calls for every agent every day. epiworldpy compiles epiworld with
  its default single-precision `float`, and with the same seed it still gives
  the same epidemics as the other two.
- **Covasim**: native disease progression restricted to SEIRH. Waning is off,
  and individual transmissibility and viral load are fixed. The severe state
  serves as the hospitalization proxy.
- **Starsim**: its built-in `ss.SEIR` disease, subclassed to add a hospital
  state, on a network module holding the shared edge list. Transmission is
  Starsim's own per-edge, per-day draw with a daily probability
  (`ss.probperday`). Durations are sampled when an agent enters a state:
  exponential latent, infectious, and hospital periods, with hospitalization
  decided by a Bernoulli draw with the lifetime probability.
- **EoN**: a continuous-time rate graph run with its event-driven
  `fast_simple_contagion` algorithm.
- **epydemic**: a `CompartmentedModel` under `SynchronousDynamics`.
- **ixa**: one plan per day, ixa's built-in contact network, and an indexed
  disease-status property. The competing infectious transitions reuse
  epiworldR's roulette rule.
- **individual**: a `CategoricalVariable` for the disease state, the built-in
  `bernoulli_process()` for E $\rightarrow$ I and H $\rightarrow$ R, and
  `fixed_probability_multinomial_process()` for leaving I, to H with the
  lifetime hospitalization probability and otherwise to R. individual has no
  contact networks, so infection is a process written for this benchmark over
  an adjacency list built from the edge list: a susceptible agent with $k$
  infectious neighbours is infected with probability $1 - (1 - \beta)^k$, as
  in epiworld. Updates queued during a day apply at its end, so the steps are
  synchronous.
- **FRED** ([PublicHealthDynamicsLab/FRED](https://github.com/PublicHealthDynamicsLab/FRED){target="_blank"}):
  a compiled simulator driven by its own model language, with no edge-list
  network primitive comparable to the other engines' graphs. The runner
  ([`run_FRED.py`](runners/run_FRED.py)) writes a synthetic population of
  single-person households (so FRED's own household mixing adds no contacts
  beyond the network) and loads the benchmark's edges verbatim as
  `Contact.add_edge` properties on FRED's `Network` group type, with
  transmission restricted to that network. The disease is a FRED condition
  with one state per compartment, and FRED's `import_count()` exposes the
  seed cases on day 0. Each day, an infectious agent draws about
  `transmissibility` × degree contacts without replacement among its
  neighbours, so every neighbour is reached with that daily probability, and
  the runner uses the same per-day mapping as the synchronous engines.
  Latent, infectious, and hospital stays are whole days with the configured
  means. FRED's `days` counts day 0, which here only places the seed cases,
  so the runner asks for one more day than the other engines run. The build
  patches an upstream off-by-one in `Date::setup_dates()` that corrupts the
  heap at large populations (see the
  [Dockerfile](../.devcontainer/Dockerfile)). FRED runs one replicate per
  process and must reread and rebuild everything each time. Its timings are
  split with FRED's own lap timers: parsing the generated model file (which
  holds every edge) and the population files counts as reading, like the
  other runners' edge-file parsing; building places, the population, and
  the network, plus the days themselves, count as simulation, as ixa's and
  Starsim's per-replicate builds do.
- **Agents.jl** ([JuliaDynamics/Agents.jl](https://github.com/JuliaDynamics/Agents.jl){target="_blank"}):
  a `StandardABM` with a per-agent status and a per-day `model_step!` that
  scans each infectious agent's adjacency list, matching the synchronous
  daily semantics of epiworldR, epydemic, and individual. The adjacency list
  and transmission math are written for this benchmark, since Agents.jl has
  no built-in epidemic model or contact network. Before any timer starts, the
  runner steps a throwaway two-agent model so that Julia compiles its
  methods; otherwise about 0.1 seconds of compilation lands in the first
  simulation.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

The simulation time above covers what an engine has to redo for every
replicate on the same network (see
[what is measured](../README.md#run-time)). *Build* is the rest of the
setup: turning the edge list into the engine's population and network, which
a user does once per network. *Build + simulate* is the time to a first
result once the edge list is in memory. Reading the edge file is shown but
left out of it, because its cost depends on each language's file parsing
rather than on the engine.

```{r}
#| label: phases
table_phases(bench)
```

### The epiworld family

The three epiworld runners build the same model on the same C++ core, and
in this scenario they produce identical epidemics for every seed. This table
compares the two wrappers with the C++ runner, replicate by replicate.

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Memory

Resident memory (RSS) of each run, in MiB, as median [Q1, Q3] across
replicates. *Simulation memory* is the peak during the simulate timer less
the memory held when it started: the extra memory one replicate needs.
*Model footprint* is what building the reusable model adds after the edge
list is read, and *baseline* is the runtime after its imports. FRED runs as
a child process, so only its overall peak is known. See "Memory" under "What
is measured" in the [overview](../README.md#memory).

```{r}
#| label: memory
table_memory(bench)
```

```{r}
#| label: memory-plot
#| fig-width: 10
#| fig-height: 9
plot_memory(bench)
```

### Epidemiological sanity checks

Speed is interpretable only if the simulations produce plausible epidemics.
These checks show the final attack rate and peak hospitalization load.
Differences also expose the engines' non-identical time semantics and
Covasim's native symptom/severe bookkeeping.

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The calibrated outcomes are closely aligned within each population size"))
}
```

### Code required to define the model

A rough measure of effort: how much code each engine needs to express the
model. The counting rules are described in the
[project overview](../README.md#measuring-implementation-effort), and the
counted regions are listed in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench)
```

Model lines vary with how much of the model an engine provides built in. For
example, EoN describes transitions as rate graphs, Covasim needs its native
parameters overridden to restrict it to SEIRH, Starsim needs a subclass of
its SEIR to add the hospital state, and the ixa runner writes its daily step
by hand, as the individual and Agents.jl runners write their infection
processes. The FRED runner writes FRED's model file and synthetic population
from Python, so its lines are the Python that generates them. ixa needs a
Cargo project that is compiled before it runs, and FRED a compiled binary.

## Interpretation

### Calibration

The shared analytic transmission mapping produced different realized attack
rates across frameworks. Calibration therefore selected size-specific factors
for restricted Covasim (0.715 at 10,000 agents and 0.710 at 100,000), ixa
(0.983 and 0.988), individual (0.951 at both), FRED (0.911 and 0.898), and
Agents.jl (0.981 and 0.982). epiworldR needs none: its
uncalibrated median attack rates match EoN's exact continuous-time results,
and so do epiworld's and epiworldpy's, which run the same model. Starsim
needs none either (uncalibrated medians 0.387 and 0.060). All 100 replicates
in each adjusted cell were then rerun. These empirical corrections are intentionally
population-specific; they align benchmark outcomes but mean that none of the
calibrated engines has a strictly analytic $R_0$ of 2. The values live in
[`scenario.toml`](scenario.toml) and participate in the cache fingerprint.

Covasim's factors were calibrated after restricting it to fixed transmission
and permanent immunity. ixa shares the synchronous daily semantics of
epiworldR and epydemic, but its uncalibrated attack rates sat above the other
engines: it seeds the initial cases as infectious and samples each infectious
contact independently rather than with epiworld's roulette. individual's
sat higher still (0.441 and 0.074). It too seeds the initial cases as
infectious, and its infectious period averages the full seven days, where the
epiworld family's competing daily hospitalization rate shortens it slightly.
Agents.jl seeds the initial cases as infectious too, and its uncalibrated
medians (0.414 and 0.065) sat slightly above the target, like ixa's. FRED's
sat far above (0.519 and 0.095). Its contact draw reaches each neighbour with
the same daily probability as the synchronous engines, but an agent that
becomes infectious at the start of a day transmits that same day, where a
synchronous engine's new case first transmits on the next step. Each
generation is therefore about a day shorter, and within 100 days the outbreak
grows further. FRED and Agents.jl were calibrated against epiworldR's
medians over 100 replicates per candidate factor, outside the timed runs.

### The language layer costs nothing measurable in the simulation

The three epiworld runners build the same model and run the same epidemics,
and they take the same time within noise at both sizes (see the table
above). Earlier versions of this report found the C++ runner slower than
the wrappers at 100,000 agents. That came from how epiworld built the
network, not from the language layer, and it went away with the
counting-sort network build
([UofUEpiBio/epiworld#274](https://github.com/UofUEpiBio/epiworld/issues/274)).

Outside the simulation, the wrappers cost something. At 100,000 agents,
building the model from the edge list takes the C++ runner
`r s3(median_of("epiworld", 1e5, "build_seconds"))` seconds, against
`r s3(median_of("epiworldR", 1e5, "build_seconds"))` for epiworldR and
`r s3(median_of("epiworldpy", 1e5, "build_seconds"))` for epiworldpy, which
first convert the edge list into their own types. Reading the edge file also
differs by language, but that is the benchmark's plumbing rather than the
engine's, so the time-to-first-result table leaves it out.

### Why the two fastest engines differ

epiworld 0.16.1 made network transmission push-based when that is cheaper:
infectious agents push infection to their neighbours instead of every queued
susceptible agent scanning all of its neighbours for infectious ones (the
change proposed in
[UofUEpiBio/epiworld#264](https://github.com/UofUEpiBio/epiworld/issues/264)).
Together with the other changes through 0.17.0, this made epiworldR about
four times faster here than 0.15.1.0 was at 100,000 agents. Both engines'
daily work now follows the outbreak rather than the population.

Per replicate, ixa is a little faster at 10,000 agents, and epiworld is
faster at 100,000: ixa takes `r ratio("ixa", "epiworld", 1e4)` and
`r ratio("ixa", "epiworld", 1e5)` times as long as the C++ runner. ixa's `execute()` can run a context only
once, so every replicate rebuilds the population, network, and index, which
grows with the population. epiworld builds its model once and resets it at
the start of every `run()`, which is about twenty times cheaper. ixa's
`execute()` alone is faster than epiworld's `run()`, because epiworld looks
up the transmission probability by name for every contact it tries and
re-initializes every agent in `reset()`. [analysis.md](../analysis.md)
measures each of these, and shows how they map onto every scenario.

### Equivalence across engines

epiworldR, epydemic, ixa, individual, and Agents.jl use synchronous daily
transitions, while EoN uses continuous-time hazards, simulated exactly. FRED
steps daily too, but applies each day's transitions before that day's
transmission, as described above. Starsim also steps
daily, but samples each agent's time in a state when it enters it, and
applies transitions before transmission within a step. Covasim retains its native
exposed, infectious, symptomatic, and severe bookkeeping, with severe
prevalence used as the hospitalization proxy. Its runner disables waning
immunity and fixes individual transmissibility and viral load to one, making
recovered people permanently removed and removing those sources of
heterogeneity. Covasim does not expose a public switch to remove the
remaining symptom/severity bookkeeping. Thus this report measures restricted,
representative framework throughput under aligned network and disease
targets; it does not claim bit-for-bit epidemiological equivalence.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
