---
title: "Scenario 00: SEIRH baseline"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_00")
```

[Back to the project overview](../README.md) ·
[Methods](../docs/methods.md) · [Results](../docs/results.md)

The reference model for the benchmark. Every later scenario adds complexity to
it, and the [results](../docs/results.md) compare each later scenario with this
one.

## Model

The [common model](../docs/methods.md#common-model) with no interventions, on
the shared Watts–Strogatz network (mean degree 10, rewiring probability 0.05)
at 10,000 and 100,000 agents, with 100 replicates per engine and size.

```{r}
#| label: parameters
table_parameters(bench)
```

## Engine notes

Every engine runs its baseline implementation, described in
[Engines](../docs/methods.md#engines). The three epiworld runners build the
same model and, in this scenario, produce identical epidemics for every seed.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 13
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by population size and replicate seed. Values above one
mean that epiworldR completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

*Build* turns the edge list into the engine's model, once per network;
*build + simulate* is the time to a first result once the edge list is in
memory ([Timing](../docs/methods.md#timing)).

```{r}
#| label: phases
table_phases(bench)
```

### Memory

Median [Q1, Q3] in MiB ([Memory](../docs/methods.md#memory)).

```{r}
#| label: memory
table_memory(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

### Model lines

Counted as described in
[Implementation size](../docs/methods.md#implementation-size), over the
regions in [`code_regions.yml`](code_regions.yml).

```{r}
#| label: code-effort
table_code_effort(bench)
```

## Notes

**Calibration.** epiworldR, its two siblings, EoN, epydemic, Starsim, and
ABM run uncalibrated: their medians already agree. The others use size-specific
factors at 10,000 and 100,000 agents (in [`scenario.toml`](scenario.toml)):
Covasim 0.715 and 0.710, ixa 0.983 and 0.988, individual 0.951 at both, FRED
0.911 and 0.898, Agents.jl 0.981 and 0.982, and EpiModel 0.948 and 0.940.
Their uncalibrated medians sat above the target: ixa's, individual's (0.441 and 0.074), and Agents.jl's
(0.414 and 0.065) because they seed the initial cases infectious, individual's
also because its infectious period averages the full seven days, where the
epiworld family's competing hospitalization rate shortens it; FRED's (0.519
and 0.095) because each of its generations is about a day shorter, and
EpiModel's (0.458 and 0.086) because its progression module runs before its
infection module, so a new infectious case transmits on the day it turns
infectious. Covasim's
were calibrated after restricting it to fixed transmission and permanent
immunity.

**Model lines.** They vary with how much of the model an engine provides.
EoN describes transitions as rate graphs, Covasim overrides its native
parameters to restrict itself to SEIRH, Starsim subclasses its SEIR for the
hospital state, the ixa, individual, and Agents.jl runners write their
daily step or infection process by hand, and the EpiModel runner writes its
infection, progression, and prevalence modules. The FRED runner's lines are the
Python that writes FRED's model file and population.
