---
title: "Scenario 03: SEIRH at 1,000,000 agents"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_03")
full <- bench$full
base <- bench$every[bench$every$scenario == "scenario_00", ]
attack <- function(data, n) {
  if (nrow(data)) sprintf("%.2g", median(data$attack_rate[data$n == n])) else "?"
}
cases <- function(data, n) {
  if (!nrow(data)) return("?")
  d <- data[data$n == n, ]
  format(round(median(d$n - d$final_susceptible), -2), big.mark = ",")
}
```

[Back to the project overview](../README.md) ·
[Methods](../docs/methods.md) · [Results](../docs/results.md) ·
[Scenario 00](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md) at 1,000,000 agents, ten times its
largest size. It asks how each engine's time grows with the population when
the outbreak does not: with 100 seed cases and the same per-contact risk,
about `r cases(base, 100000)` agents are infected at 100,000 agents and
`r cases(full, 1000000)` at 1,000,000 (median attack rates
`r attack(base, 100000)` and `r attack(full, 1000000)`). The
[results](../docs/results.md#scaling) compare the growth across engines.

## Model

The model, parameters, network settings, seed cases, and runners
([`runners/`](runners)) are scenario 00's. Each engine runs 20 replicates
rather than 100, because the slowest take tens of seconds per replicate, and
the `[design]` table in [`scenario.toml`](scenario.toml) keeps this scenario
sequential whatever `N_THREADS` says
([why](../docs/methods.md#execution-protocol)).

```{r}
#| label: parameters
table_parameters(bench)
```

## Engine notes

None: every runner is scenario 00's.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

```{r}
#| label: simulation-time-plot
#| fig-width: 10
#| fig-height: 7
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by replicate seed. Values above one mean that epiworldR
completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

[Timing](../docs/methods.md#timing) defines build and simulate.

```{r}
#| label: phases
table_phases(bench)
```

### Memory

Median [Q1, Q3] in MiB ([Memory](../docs/methods.md#memory)).

```{r}
#| label: memory
table_memory(bench)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

### Model lines

The runners are scenario 00's, so are their model lines (see
[scenario 00](../scenario_00/README.md#model-lines)).

## Notes

**Calibration.** The factors were not recalibrated at this size. Each engine
reuses its 100,000-agent factor from scenario 00, which barely moved between
10,000 and 100,000 agents, and the attack rates above stay aligned.

**Memory sets the limits here.** At this size some engines need several
gigabytes per replicate (see the memory table), which is why four concurrent
replicates competed for memory and why this scenario runs one at a time.
