---
title: "Scenario 03: SEIRH baseline at 1,000,000 agents"
subtitle: "Network epidemic ABM speed benchmark"
date: last-modified
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
bench <- load_benchmark("scenario_03")
baseline <- "scenario_00"
full <- bench$full
base <- bench$every[bench$every$scenario == baseline, ]
median_of <- function(data, engine, n, column = "simulate_seconds") {
  d <- data[data$engine == engine & data$n == n, ]
  if (nrow(d)) median(d[[column]]) else NA
}
seconds <- function(x) if (is.na(x)) "?" else sprintf("%.2g", x)
growth <- function(engine) {
  sprintf("%.1f", median_of(full, engine, 1e6) / median_of(base, engine, 1e5))
}
build <- function(engine) {
  d <- full[full$engine == engine, ]
  if (nrow(d)) sprintf("%.2g", median(d$setup_seconds - d$read_seconds)) else "?"
}
attack <- function(data, n) {
  if (nrow(data)) sprintf("%.2g", median(data$attack_rate[data$n == n])) else "?"
}
cases <- function(data, n) {
  if (!nrow(data)) return("?")
  d <- data[data$n == n, ]
  format(round(median(d$n - d$final_susceptible), -2), big.mark = ",")
}
```

[Back to the project overview](../README.md) ·
[Scenario 00 report](../scenario_00/README.md)

[Scenario 00](../scenario_00/README.md) at 1,000,000 agents. The model, the
parameters, the network's mean degree and rewiring probability, the 100 seed
cases, and every runner are those of scenario 00; only the population is ten
times its largest size. Each engine runs 20 replicates rather than 100,
because the slowest ones take several seconds per replicate here, and one
replicate at a time rather than four ([see why](#why-this-scenario-runs-one-replicate-at-a-time)).
The size, replicate count, and concurrency are set in the `[design]` table of
[`scenario.toml`](scenario.toml).

The question this scenario asks is how each engine's time grows with the
population when the outbreak does not. With 100 seed cases and the same
per-contact risk, the epidemic stays about the same size in absolute terms:
about `r cases(base, 100000)` agents are infected at 100,000 agents and
`r cases(full, 1000000)` at 1,000,000 (median attack rates
`r attack(base, 100000)` and `r attack(full, 1000000)`). An engine whose
work follows the outbreak should barely slow down. An engine that touches
every agent every day should slow down about tenfold.

## Model

See [scenario 00](../scenario_00/README.md#model) for the model, its
parameters, the timing conventions, and each engine's implementation. The
runners in [`runners/`](runners) are copies of scenario 00's.

The transmission multipliers were not recalibrated at this size. Each engine
reuses its 100,000-agent factor from scenario 00, which barely moved between
10,000 and 100,000 agents. The attack rates below show that the engines stay
aligned.

## Results

```{r}
#| label: completion-status
#| results: asis
completion_status(bench)
```

```{r}
#| label: environment
table_environment(bench)
```

```{r}
#| label: summary-table
table_summary(bench)
```

### Simulation time

The primary measure is wall-clock time inside each engine's simulation call.
The logarithmic scale keeps fast and slow engines legible in one panel.

```{r}
#| label: simulation-time-plot
#| fig-width: 12
#| fig-height: 5
if (bench$has_full) print(plot_simulation_time(bench))
```

### Speed relative to epiworldR

Ratios are matched by replicate seed. Values above one mean that epiworldR
completed the simulation call faster.

```{r}
#| label: speedup
table_speedup(bench)
```

### Time to a first result

The simulation time above covers what an engine has to redo for every
replicate on the same network (see
[what is measured](../README.md#run-time)). *Build* is the rest of the
setup: turning the edge list into the engine's population and network, which
a user does once per network. *Build + simulate* is the time to a first
result once the edge list is in memory. Reading the edge file is shown but
left out of it, because its cost depends on each language's file parsing
rather than on the engine.

```{r}
#| label: phases
table_phases(bench)
```

### The epiworld family

```{r}
#| label: epiworld-family
table_speedup(bench, reference = "epiworld", engines = c("epiworldR", "epiworldpy"))
```

### Growth with the population

Median simulation time at each size, the first two from scenario 00. The
growth column divides the time at 1,000,000 agents by the time at 100,000;
the population grew tenfold, the outbreak hardly at all. The last column is
the time to build the model from the edge list, which is not part of the
simulation time (see [Building once](#building-once)).

```{r}
#| label: scaling
table_scaling(bench, baseline)
```

### Epidemiological sanity checks

```{r}
#| label: outcomes-table
table_outcomes(bench)
```

```{r}
#| label: outcomes-plot
#| fig-width: 9
#| fig-height: 5
if (bench$has_full) {
  print(plot_outcomes(bench, "The outcomes stay aligned at 1,000,000 agents without recalibration"))
}
```

## Interpretation

### How time grows with the population

The growth factors compare this scenario, run one replicate at a time, with
scenario 00 at 100,000 agents, run four at a time. Concurrency slows some
engines more than others, so the factors are approximate, and slightly low
for the engines it slows most.

- **The epiworld runners** grow by `r growth("epiworld")` (C++),
  `r growth("epiworldR")` (epiworldR), and `r growth("epiworldpy")`
  (epiworldpy) times. Their daily work follows the outbreak, which barely
  grew. What does grow is `reset()` at the start of `run()`, which
  re-initializes every agent; at this size it is nearly half of the run.
- **individual** grows by `r growth("individual")` times. Its infection
  process visits the infectious agents' neighbours, but it also tabulates
  contacts and intersects bitsets over the whole population every day.
- **ixa** grows by `r growth("ixa")` times. Its `execute()` follows the
  outbreak, but every replicate first rebuilds the context (a million
  entities, five million edges, and the index), which grows with the
  population and is now most of its time.
- **EoN** (`r growth("EoN")` times) and **epydemic** (`r growth("epydemic")`)
  follow the outbreak in their event handling, but set up state for every
  node of the network, in Python, at the start of each run.
- **Covasim** (`r growth("covasim")` times) and **Starsim**
  (`r growth("starsim")`) update every agent's arrays every day, and Starsim
  also draws transmission on every edge, so they grow about as fast as the
  population.

At 1,000,000 agents, epydemic takes about
`r seconds(median_of(full, "epydemic", 1e6))` seconds per replicate and
Starsim `r seconds(median_of(full, "starsim", 1e6))`, against
`r seconds(median_of(full, "epiworld", 1e6))` for the C++ runner and
`r seconds(median_of(full, "ixa", 1e6))` for ixa.
[analysis.md](../analysis.md) measures where epiworld's and ixa's time goes.

### Building once

Building the model from the edge list, which a user does once per network,
takes the epiworld runners `r build("epiworld")` (C++), `r build("epiworldR")`
(epiworldR), and `r build("epiworldpy")` (epiworldpy) seconds: several times
longer than a 100-day run, but paid once however many replicates follow. EoN
spends `r build("EoN")` seconds and epydemic `r build("epydemic")` building
their NetworkX graphs, which they also reuse. ixa and Starsim have
almost nothing to build once, because they rebuild their whole model for
every replicate, and that time is in their simulation time above.

### Why this scenario runs one replicate at a time

With four workers, the scheduler ran four replicates of the same engine at
once, and at this size they competed for memory: epydemic took 57 to 67
seconds per replicate, against about 17 run alone, and epiworldpy and EoN
slowed too. The `workers = 1` setting in [`scenario.toml`](scenario.toml)
makes this scenario run sequentially whatever `N_THREADS` says. Its times are
therefore not directly comparable with the other scenarios', which run four
replicates at a time.

## Recorded versions

```{r}
#| label: versions
table_versions(bench)
```
