---
title: "Methods"
subtitle: "Network epidemic ABM speed benchmark"
format:
  gfm:
    toc: true
bibliography: ../references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
all <- load_all()
```

[Back to the project overview](../README.md) ·
[Results](results.md)

This document describes how the benchmark is built and run: the engines,
the common model, the networks, calibration, the execution protocol, and
what each run measures. The [results](results.md) compare the engines across
scenarios, and each scenario's report covers what that scenario adds.

## Engines

Thirteen engines run every scenario ([Engines](engines.md) describes what each
is). Each runner is a short program in the
engine's own language that reads the shared edge list, builds the model with
the engine's own tools, runs one replicate, and writes one result record.

```{r}
#| label: engines
engines <- data.frame(
  engine = c("epiworld", "epiworldR", "epiworldpy", "covasim", "starsim", "EoN",
             "epydemic", "ixa", "individual", "ABM", "EpiModel", "FRED", "Agents.jl"),
  name = c("epiworld [@vegayonEpiworld2026]", "epiworldR [@meyerEpiworldRFastAgentBased2023]",
           "epiworldpy [@vegayonEpiworldpy2026]", "Covasim [@kerrCovasimAgentbasedModel2021]",
           "Starsim [@kerrStarsim2025]", "EoN* [@millerEoNEpidemicsNetworks2019]",
           "epydemic [@dobson2022epydemic]", "ixa [@ixa2026]",
           "individual [@charlesIndividual2021]", "ABM* [@maABM2025]",
           "EpiModel [@jennessEpiModel2018]", "FRED [@grefenstetteFRED2013]",
           "Agents.jl [@datserisAgentsjl2024]"),
  language = c("C++", "R (C++ core)", "Python (C++ core)", "Python", "Python", "Python",
               "Python", "Rust", "R (C++ core)", "R (C++ core)", "R", "C++, driven by its own model language from Python",
               "Julia"),
  features = c(
    "Custom model from `add_state()` and the update-function factories",
    "The same model through the R wrapper (`update_fun_susceptible()`, `update_fun_rate()`)",
    "The same model through the Python wrapper (`UpdateFun.susceptible()`, `UpdateFun.rate()`)",
    "Native disease progression, restricted to SEIRH; severe state as the hospitalization proxy",
    "Built-in `ss.SEIR`, subclassed to add a hospital state; network module over the edge list",
    "Rate graphs run with the event-driven `fast_simple_contagion`",
    "`CompartmentedModel` under `SynchronousDynamics`",
    "One plan per day, the built-in contact network, an indexed disease-status property",
    "`CategoricalVariable` and built-in transition processes; infection process written for the benchmark",
    "`Simulation` with exponential transitions over a `Contact` subclass serving the edge list",
    "`netsim()` with SEIRH modules on a fixed network: no ERGM fit, the edge list installed at initialization",
    "A FRED condition per compartment over a `Network` group type loaded with the edges",
    "`StandardABM` with a daily `model_step!` over an adjacency list written for the benchmark"),
  time = c("Synchronous daily", "Synchronous daily", "Synchronous daily",
           "Daily, native", "Daily; durations sampled on entry", "Continuous time (exact)",
           "Synchronous daily", "Synchronous daily", "Synchronous daily",
           "Continuous time; counts reported daily", "Daily; progression, then transmission",
           "Daily; transitions before transmission", "Synchronous daily")
)
versions <- unique(all$rows[c("engine", "engine_version")])
engines$version <- versions$engine_version[match(engines$engine, versions$engine)]
knitr::kable(engines[c("name", "language", "version", "features", "time")],
  col.names = c("Engine", "Language", "Recorded version", "Model as implemented", "Time"),
  row.names = FALSE)
```

\* A continuous-time engine: it simulates events at exact times, and the
benchmark records the daily totals. The same asterisk marks it in every
figure and table.

The recorded version is the one each result record carries. Versions built
from a pinned commit end in that commit (`+g…`), so moving a pin
invalidates the engine's published results.

Implementation details that matter for interpreting results:

- **The epiworld family.** The three runners build the same model on the same
  C++ core, so their differences in the simulation come from the language
  layer. The C++ runner is compiled with the flags R uses for epiworldR
  (`-O2 -DNDEBUG`, double precision); epiworldpy compiles epiworld with its
  default single-precision `float`. epiworldpy's `UpdateFun` factories were
  added for this benchmark
  ([UofUEpiBio/epiworldpy#17](https://github.com/UofUEpiBio/epiworldpy/pull/17)).
  Before, a custom model's states needed Python callbacks for every agent
  every day.
- **Covasim** runs with waning off and individual transmissibility and viral
  load fixed to one, so recovered people are permanently removed. Its
  remaining symptom and severity bookkeeping has no public switch.
- **Starsim** draws transmission on every edge every day with a daily
  probability (`ss.probperday`), and samples exponential latent, infectious,
  and hospital durations when an agent enters a state.
- **ixa**'s competing infectious transitions reuse epiworld's roulette rule;
  it samples each infectious contact independently.
- **individual** has no contact networks. Its infection process, written for
  the benchmark over an adjacency list, infects a susceptible agent with $k$
  infectious neighbours with probability $1 - (1 - \beta)^k$, as epiworld
  does. Updates queued during a day apply at its end.
- **ABM** is event-driven: every transition fires after an exponential
  waiting time, so it uses EoN's mapping from $R_0$ to a per-contact rate
  rather than the daily probabilities of the synchronous engines. Its
  built-in networks are random mixing and a configuration model sampled from a
  degree distribution, neither of which can load an edge list. The runner
  writes a `Contact` subclass whose `contact()` returns an agent's neighbours
  from an adjacency list, which ABM then treats as a per-neighbour
  transmission process. The final compartments are counted from the agents'
  states after the simulate timer stops, and `peak_hospitalized` is the
  largest end-of-day count of an ABM counter, not the exact continuous-time
  peak. In version 0.4.3, an initial state given as a character string makes
  ABM build a temporary list that its garbage collection handling can lose,
  which corrupts one agent now and then
  ([junlingm/ABM#4](https://github.com/junlingm/ABM/issues/4)), so the runner hands ABM states that
  are already lists. The scenario 01 vaccine puts protected agents in a state
  that no transition touches, and the scenario 02 outputs come from a
  callback on the infection transition and from counters on each pair of
  states.
- **EpiModel** draws its networks from a fitted exponential random graph
  model. The benchmark's network is fixed, so the runner skips both the fit
  and the draw: the formation model is an edges term held at the shared
  network's density (`~offset(edges)`, which leaves ergm nothing to estimate),
  a replacement initialization module installs the edge list where EpiModel
  would put its first draw, and the network-resimulation module does nothing.
  EpiModel keeps the network as an edge list (`tergmLite = TRUE`), its form
  for large models. Its default, a `networkDynamic` object that records the
  network's history, was 30 to 40 times slower on the same fixed network in a
  side test at 10,000 and 100,000 agents, because every day it also reads the
  edges out of that object and writes every agent's status into it. The SEIRH
  disease is written as modules, as in the EpiModel Gallery: infection over
  `discord_edgelist()` with probability $1 - (1 - \beta)^k$ for $k$ infectious
  neighbours, as epiworld does, and progression with daily probabilities.
  EpiModel runs user modules before its built-in ones, so each day
  progression comes first and an agent that becomes infectious transmits the
  same day. `discord_edgelist()` shuffles and scans the whole edge list every
  day, whatever the prevalence, and it is most of EpiModel's run time. The
  scenario 01 vaccine is a protection attribute that the infection module
  skips, and the scenario 02 outputs are EpiModel's transmission matrix and
  epidemic statistics for each flow.
- **FRED** has no edge-list network primitive. Its runner writes a synthetic
  population of single-person households, so FRED's household mixing adds no
  contacts, and loads the edges verbatim as `Contact.add_edge` properties on
  FRED's `Network` group type, with transmission restricted to it. Each day an
  infectious agent draws about `transmissibility` × degree contacts without
  replacement among its neighbours, so each neighbour is reached with the
  per-day probability of the synchronous engines. Stays are whole days with
  the configured means. FRED's `days` counts day 0, which only places the seed
  cases, so the runner asks for one more day. The image patches an upstream
  off-by-one in `Date::setup_dates()` that corrupts the heap at large
  populations (see the [Dockerfile](../.devcontainer/Dockerfile)). An agent
  that becomes infectious at the start of a day transmits that same day,
  where a synchronous engine's new case first transmits on the next step.
- **Agents.jl** has no built-in epidemic model or contact network; the
  adjacency list and transmission are written for the benchmark. Before any
  timer starts, the runner steps a throwaway two-agent model so that Julia
  compiles its methods; otherwise about 0.1 seconds of compilation would land
  in the first simulation.

### Inclusion and exclusion

The engines included so far share what the common design needs: each runs
the common model over an arbitrary undirected edge list, from a script, with
a seed, and reports the five compartment counts. MEmilio and Epiabm were
assessed and excluded for now: neither takes an edge list as its contact
structure, and both have disease models with no simple mapping onto the
common SEIRH model. [Engines considered and not included](excluded-engines.md)
states the inclusion criteria, the reasons for each exclusion, and the
candidates not yet assessed.

## Common model

Every scenario starts from the same SEIRH model: susceptible →
exposed → infectious → recovered, with a competing infectious →
hospitalized → recovered branch. Its parameters are in each scenario's
`scenario.toml`:

| Parameter | Value | Meaning |
|:----------|------:|:--------|
| `target_r0` | 2.0 | Early-outbreak $R_0$ used by the transmission mapping |
| `initial_infected` | 100 | Seed cases (10 in the smoke profile) |
| `latent_days` | 4.0 | Mean exposed period |
| `infectious_days` | 7.0 | Mean infectious period |
| `hospitalization_probability` | 0.05 | Lifetime probability that an infectious agent is hospitalized |
| `hospital_days` | 7.0 | Mean hospital stay |

**Transmission mapping.** With mean degree $k$ and recovery rate
$\gamma = 1/\mathrm{infectious\_days}$, every engine starts from the per-contact
transmissibility $T = \min(0.999,\ R_0 / \max(1, k - 1))$. The daily engines
convert it into a per-day, per-edge probability through geometric competing
risks, $\beta = T\gamma / (1 - T(1 - \gamma))$, and the daily hospitalization
probability likewise, $h = p\gamma / (1 - p(1 - \gamma))$. EoN, in continuous
time, uses the rate $\tau = T\gamma / (1 - T)$ and hospitalization rate
$p\gamma / (1 - p)$. The mapping divides by the mean degree minus one, which
is exact only for a network with near-constant degree; scenario 04's report
discusses what that means on its heterogeneous network. A scenario's
calibration factors multiply $\beta$ (see [Calibration](#calibration)).

**Seed cases.** epiworldR's API seeds a custom model's initial cases only in
the state new infections enter, so they start exposed there, and the C++ and
Python runners do the same so that all three build exactly the same model.
Starsim's SEIR and FRED's import also start seed cases exposed. Every other
engine seeds them infectious.

**Equivalence.** The engines share the network, the parameters, and the
transmission mapping, not their time semantics: synchronous daily steps,
continuous time, sampled durations, or Covasim's native bookkeeping. The
benchmark therefore measures representative throughput under aligned
network and disease targets, checked by the outcomes in every report, and
does not claim bit-for-bit epidemiological equivalence.

## Networks

**Watts–Strogatz.** Scenarios 00 to 03 use a Watts–Strogatz graph with mean
degree 10 and rewiring probability 0.05 (`[network]` in `config.toml`),
generated once per population size with NetworkX and seed 20260907 plus the
size's index. Mean degree is held fixed rather than graph density, which
keeps the edge count, run time, and memory from growing quadratically with
the population.

**GeoPops.** Scenario 04 uses a contact network collapsed from GeoPops'
Spartanburg County, South Carolina synthetic population: four contact layers
(household, workplace, school, group quarters) restricted to the first
200,000 people, unioned into one undirected, unweighted graph, with people
who have no contacts among them dropped. Its report gives the construction,
provenance, and the resulting graph's structure.

Every engine reads the same gzipped, tab-separated edge list at a given size
and treats it as undirected. A network's identity is the SHA-256 of its
decompressed edge list, so it does not depend on the compressor; it is part
of every replicate's fingerprint.

## Calibration

The shared mapping gives different realized attack rates in engines with
different time semantics. Calibration aligns them with one protocol:

1. Run every engine uncalibrated at each population size of the scenario,
   with the full replicate count.
2. Compare each engine's median final attack rate with epiworldR's, whose
   uncalibrated medians match EoN's exact continuous-time results.
3. For an engine that differs, choose a size-specific multiplier on $\beta$
   (`<engine>_transmission_multiplier_<n>` in `scenario.toml`) whose median
   attack rate over 100 replicates per candidate factor matches epiworldR's.
   Candidate runs are separate from the timed runs.
4. Rerun every replicate of each adjusted cell with its factor.

The factors are empirical and size-specific: they align outcomes, but no
calibrated engine has a strictly analytic $R_0$ of 2. They are part of every
replicate's fingerprint. Each scenario's report states which factors it
uses and why.

## Execution protocol

- **Container.** Published results run inside the image built from
  [`.devcontainer/Dockerfile`](../.devcontainer/Dockerfile), which pins every
  engine and toolchain (see [Environment](#environment)).
- **One replicate at a time.** Replicates run sequentially, each in a fresh
  process, so no state leaks between replicates and a failure affects only
  its own record. Native math libraries are pinned to one thread
  (`OMP_NUM_THREADS`, `OPENBLAS_NUM_THREADS`, `MKL_NUM_THREADS`,
  `VECLIB_MAXIMUM_THREADS`, `NUMEXPR_NUM_THREADS`, `NUMBA_NUM_THREADS`, and
  `RCPP_PARALLEL_NUM_THREADS` set to 1), and `PYTHONHASHSEED` is fixed.
- **Seeds.** Replicate $r$ at the population size with index $i$ uses seed
  $8675309 + 100000\,i + r$. Seeds do not depend on the scenario, so two
  scenarios can be compared replicate by replicate.
- **Replicates.** 100 replicates of 100 days per engine and size, at 10,000
  and 100,000 agents (`[study]` in `config.toml`). A scenario's `[design]`
  table can replace the sizes and the count: scenarios 03 and 04 run 20
  replicates, at 1,000,000 and 165,865 agents.
- **Excluded from timing.** Interpreter and package startup, network
  generation, and Julia's compilation, which the warm-up absorbs.

**Why sequential.** Replicates running side by side contend for memory
bandwidth and, on hybrid CPUs, for performance cores, and the penalty
differs by engine, which biases the comparison rather than adding noise. In
the container (a podman VM with 10 vCPUs on an M3 Pro), a probe ran 24
replicates of each engine one, two, and four at a time at 100,000 agents,
the way the benchmark schedules them. Median `simulate_seconds` relative to
one at a time:

| Engine | Scenario 00, 2 workers | Scenario 00, 4 workers | Scenario 01, 2 workers | Scenario 01, 4 workers |
|:---|---:|---:|---:|---:|
| epiworld (C++) | 1.07 | 1.26 | 1.14 | 1.20 |
| epiworldR | 1.00 | 1.06 | 1.11 | 1.56 |
| ixa | 1.30 | 1.08 | 1.25 | 1.22 |
| Covasim | 1.00 | 1.07 | 1.06 | 1.11 |
| epydemic | 1.05 | 1.12 | 1.15 | 1.69 |

Even two workers slowed some engines by up to 30%, unevenly. A 6-worker run
of the full design inflated epiworldR's median at 100,000 agents about 2.8
times and ixa's 1.2 to 1.4 times. At 1,000,000 agents, four concurrent
processes also competed for memory and more than tripled epydemic's time (57 to 67
seconds per replicate, against about 17 alone). An earlier native probe on
the same machine (11 cores: 5 performance, 6 efficiency) found up to three
workers free but six to eight inflating times 1.1 to 2.2 times, the fastest
engine the most. Epidemiological outputs are unaffected; only timings are.

## Measures

### Timing

Each runner times its own phases:

- `simulate_seconds`, the primary measure: everything the engine has to redo
  to produce another replicate on the same network. That is its simulation
  call, including any initialization done inside it, plus whatever cannot be
  reused between runs:
  - setting up each run's initial conditions (seeding the infections and,
    where there is one, distributing a vaccine), which differ from run to
    run and which epiworld does inside `run()`; EoN's timer covers building
    its initial status dictionary and individual's its state variable;
  - for ixa, building the context (population, network, and index), because
    `execute()` runs a context only once;
  - for Starsim, building the disease, network, and `Sim` objects and
    `sim.init()`, because a `Sim` also runs only once;
  - for FRED, which runs one replicate per process, building its places,
    population, and network.

  epiworld's `run()` re-initializes its whole population for the same reason,
  so every engine pays for starting a fresh run. Bookkeeping an engine does
  while it simulates is included; reading results out of the engine
  afterwards is not (scenario 02 times it separately as `extract_seconds`).
- `read_seconds`: reading and parsing the shared edge list, which depends on
  each language's file handling rather than on the engine. FRED's parsing of
  its generated model file (which holds every edge) and population files,
  taken from FRED's own lap timers, counts here.
- `setup_seconds`: reading plus building what the engine reuses across
  replicates, such as epiworld's model or EoN's graph. Reports show the
  building part, `setup_seconds - read_seconds`, as *build*, and
  build + simulate as the time to a first result once the edge list is in
  memory.
- `total_seconds`: setup plus simulation inside the runner process.

### Memory

Each runner reads its resident memory (RSS, from `/proc/self/status`) at the
same boundaries as the timings:

- `rss_baseline_bytes`: after the runtime and the engine are loaded, before
  the edge list is read;
- `rss_after_read_bytes`: after the edge list is parsed;
- `rss_after_setup_bytes`: after the reusable model is built, just before
  the simulate timer starts;
- `peak_rss_setup_bytes`: the high-water mark (`VmHWM`) at that point;
- `peak_rss_simulate_bytes`: the high-water mark at the end of the simulate
  timer. The runner resets it (by writing `5` to `/proc/self/clear_refs`)
  just before the timer starts, so it is the peak of the simulation alone;
- `process_peak_rss_bytes`: the runner's `ru_maxrss` when it exits. On Linux,
  `ru_maxrss` also counts the memory the parent held when it forked the
  runner, so the orchestrator starts each runner through GNU `time`, whose
  own image is under a megabyte. The kernel derives `ru_maxrss` from the same
  high-water mark, so the reset lowers it too: it checks the simulation peak
  and anything after it, not the whole run.

From these, *simulation memory* (`peak_rss_simulate - rss_after_setup`) is
the extra memory one more replicate needs; *model footprint*
(`rss_after_setup - rss_after_read`) is what the reusable model holds; and
*overall peak* is the largest of `peak_rss_setup`, `peak_rss_simulate`, and
`process_peak_rss`. The phases follow the timings, so an engine that builds
its model inside the simulate timer (ixa, Starsim) shows that model as
simulation memory. R, Python, and Julia count garbage they have not collected
yet as resident; no collection is forced before a reading, as that would
change the timings. FRED runs as a child process of its Python runner, so
only its whole-run peak is known, recorded as `peak_rss_simulate_bytes`.

### Implementation size

*Model lines* counts the non-blank, non-comment lines that build the
population and network, define the disease and any interventions, and run
the simulation. Argument parsing, edge-file reading, timing, and writing
results are excluded, because every runner shares them. Each scenario's
`code_regions.yml` lists the counted regions, and the counts are recomputed
from the runner sources at every render. Each engine uses its own native
tools, so the counts reflect what a user of that engine would write. They
measure implementation size, not developer effort or usability.

## Environment

Each run that executes a scenario's replicates records where it ran in
`results/environments/<scenario>.json`; [setup.md](../setup.md#environment-records)
lists the fields. The published results:

```{r}
#| label: environment
table_environments()
```

The table describes the run that produced every engine except ABM and
EpiModel, which were added later and ran separately on the same machine and
virtual machine, one replicate at a time: ABM in the published image with
ABM installed into a mounted library, and EpiModel in an image built from the
Dockerfile with EpiModel added. Their records carry their own
`environment_id`, so each scenario report warns that its replicates mix
environments.

## Scenario ladder

```{r}
#| label: ladder
ladder <- do.call(rbind, lapply(names(all$info), function(s) {
  d <- all$design[all$design$scenario == s, ]
  data.frame(
    scenario = sprintf("[%s](../%s/README.md)", scenario_label(s), s),
    description = all$info[[s]]$description,
    agents = paste(format(d$n, big.mark = ",", scientific = FALSE, trim = TRUE), collapse = ", "),
    replicates = paste(unique(d$replicates), collapse = ", ")
  )
}))
knitr::kable(ladder, col.names = c("Scenario", "What it adds", "Agents", "Replicates per engine and size"),
             row.names = FALSE)
```

## Reproducibility

The image is about 4.8 GB. Built by CI, it is published as
`ghcr.io/uofuepibio/epiworld-benchmark` and tagged with a hash of its inputs;
[setup.md](../setup.md) covers building or pulling it, and running the
benchmark natively. From a clone:

```sh
make container-image       # or pull the published image (see setup.md)
make container-check       # tests, runner builds, R package check
make container-smoke       # one tiny replicate of every engine in every scenario
make container-benchmark   # the full design
make container-report      # the overview, these documents, and every scenario report
```

The full design is 7,040 replicates. The published run took 3 h 12 min, one
replicate at a time, on an Apple M3 Pro under podman. Every replicate's
record is committed under `results/runs/` (about 14 MB), so a clone skips
what is already published; the generated networks and GeoPops sources in
`cache/` take about 33 MB and are rebuilt on demand.

## References
