---
title: "Results"
subtitle: "Network epidemic ABM speed benchmark"
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("../report/common.R")
all <- load_all()
med <- function(engine, scenario, n, column = "simulate_seconds") {
  cell_median(all, engine, scenario, n, column)
}
secs <- function(x) formatC(x, digits = 2, format = "fg")
mib <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
times <- function(x) sprintf("%.1f", x)
max_family_ratio <- local({
  rows <- all$rows
  a <- rows[rows$engine %in% c("epiworldR", "epiworldpy"), c("engine", "scenario", "n", "seed", "simulate_seconds")]
  b <- rows[rows$engine == "epiworld", c("scenario", "n", "seed", "simulate_seconds")]
  p <- merge(a, b, by = c("scenario", "n", "seed"), suffixes = c("", "_cpp"))
  max(aggregate(simulate_seconds / simulate_seconds_cpp ~ engine + scenario + n, p, median)[[4]])
})
growth <- growth_factors(all)
attack <- function(scenario, n) {
  sprintf("%.3f", median(all$rows$attack_rate[all$rows$engine == "epiworldR" &
                                                all$rows$scenario == scenario & all$rows$n == n]))
}
```

[Back to the project overview](../README.md) ·
[Methods](methods.md)

The results across every scenario, from `results/results.csv`. The
[methods](methods.md) define each measure; each scenario's report has its
full tables, including interquartile ranges.

## Run time

### Where the time goes

The median time of each phase of a replicate: reading the shared edge list,
building what the engine reuses across replicates, and simulating, which
includes everything the engine redoes for each replicate (see
[Timing](methods.md#timing)). Panels are scenarios at each population size.

```{r}
#| label: decomposition
#| fig-width: 12
#| fig-height: 9
plot_runtime_decomposition(all)
```

Simulation time is the measure every report ranks engines by. Its median in
seconds, in every scenario and size:

```{r}
#| label: simulation-time
table_by_cell(all, "simulate_seconds", 4)
```

### The language layer

The three epiworld runners build the same model on the same C++ core. The
wrappers' simulation time relative to the C++ runner, matched by seed:

```{r}
#| label: family
table_family_ratios(all)
```

In the simulation itself the language layer costs nothing measurable:
neither wrapper is ever more than
`r sprintf("%.0f%%", 100 * (max_family_ratio - 1))` slower than the C++
runner, and epiworldpy is often faster. Where the C++ runner is slowest, in
scenarios 01 and 02 at 100,000 agents, the cause is memory allocation in the
C++ runner, not the wrappers (see scenario 01's report). Outside the
simulation the wrappers cost something: at 100,000 agents, building the
model takes the C++ runner `r secs(med("epiworld", "scenario_00", 1e5, "build_seconds"))`
seconds, against `r secs(med("epiworldR", "scenario_00", 1e5, "build_seconds"))`
for epiworldR and `r secs(med("epiworldpy", "scenario_00", 1e5, "build_seconds"))`
for epiworldpy, which first convert the edge list into their own types.

### Why the two fastest engines differ

In scenarios 00 to 02, epiworld and ixa are the two fastest engines, and
epiworld is faster per replicate: in scenario 00, ixa takes
`r times(med("ixa", "scenario_00", 1e4) / med("epiworld", "scenario_00", 1e4))`
and `r times(med("ixa", "scenario_00", 1e5) / med("epiworld", "scenario_00", 1e5))`
times as long as the C++ runner at 10,000 and 100,000 agents. Both engines'
daily work follows the outbreak, through push-based transmission in epiworld
(since 0.16.1, with the push/pull choice revised in 0.17.1,
[UofUEpiBio/epiworld#281](https://github.com/UofUEpiBio/epiworld/pull/281)),
and their epidemic computations are close. The difference is what each
redoes per replicate. ixa's `execute()` runs a context only once, so every
replicate rebuilds the population, network, and index, which grows with the
population. epiworld builds its model once and re-initializes every agent at
the start of every `run()`, about twenty times cheaper; with a vaccine, it
also clones the tool for every agent it protects. Bookkeeping does not
separate them: epiworld records its outputs in every scenario, and ixa adds
the same records in scenario 02 at little cost.
[The epiworld and ixa note](epiworld-ixa.md) measures each of these.

The order changes elsewhere. At 1,000,000 agents (scenario 03), individual
is faster than ixa. On scenario 04's network, Agents.jl, individual, ixa, and
the epiworld family finish within about 20% of each other
(`r secs(med("Agents.jl", "scenario_04", 165865))` to
`r secs(med("epiworldR", "scenario_04", 165865))` seconds). Its outbreak
infects `r attack("scenario_04", 165865)` of the population, against
`r attack("scenario_00", 1e5)` in scenario 00 at 100,000 agents, so the
engines whose work follows the outbreak do far more of it.

## Scaling

The scenario 00 model at 10,000 and 100,000 agents and, in scenario 03, at
1,000,000, on the same kind of network. With 100 seed cases and the same
per-contact risk, the outbreak stays about the same size in absolute terms,
so an engine whose work follows the outbreak should barely slow down, and one
that touches every agent every day should slow down about tenfold per
decade.

```{r}
#| label: scaling
#| fig-width: 8
#| fig-height: 5
plot_scaling(all)
```

From 100,000 to 1,000,000 agents:

- **The epiworld runners** grow by `r times(growth[["epiworld"]])` (C++),
  `r times(growth[["epiworldR"]])` (epiworldR), and
  `r times(growth[["epiworldpy"]])` (epiworldpy) times. Their daily work
  follows the outbreak; what grows is `reset()`, which re-initializes every
  agent at the start of `run()`.
- **individual** grows by `r times(growth[["individual"]])` times. Its
  infection process visits the infectious agents' neighbours, but it also
  tabulates contacts and intersects bitsets over the whole population every
  day.
- **EoN** (`r times(growth[["EoN"]])` times) and **epydemic**
  (`r times(growth[["epydemic"]])`) follow the outbreak in their event
  handling, but set up state for every node, in Python, at the start of each
  run.
- **Covasim** (`r times(growth[["covasim"]])` times) and **Starsim**
  (`r times(growth[["starsim"]])`) update every agent's arrays every day, and
  Starsim also draws transmission on every edge, so they grow about as fast
  as the population.
- **ixa** grows by `r times(growth[["ixa"]])` times: its per-replicate
  rebuild of a million entities, five million edges, and the index is most of
  its time at this size.
- **FRED** grows by `r times(growth[["FRED"]])` times, faster than the
  population: it rebuilds its places, population, and network for every
  replicate.
- **Agents.jl** grows by `r times(growth[["Agents.jl"]])` times. Its daily
  step visits only the infectious agents' neighbours, but it loops over every
  agent to find them and counts the hospitalized agents over the whole
  population every day.

Building the model once from the edge list, which a user pays once per
network however many replicates follow, takes the C++ runner
`r secs(med("epiworld", "scenario_03", 1e6, "build_seconds"))` seconds at
1,000,000 agents, several times one of its replicates, and EoN and epydemic
`r secs(med("EoN", "scenario_03", 1e6, "build_seconds"))` and
`r secs(med("epydemic", "scenario_03", 1e6, "build_seconds"))` seconds for
their NetworkX graphs.

## Cost of added features

Scenario 01 adds a vaccine to scenario 00, and scenario 02 adds four outputs
to scenario 01. The median ratio of each replicate's simulation time to the
one with the same engine, size, and seed in the simpler scenario:

```{r}
#| label: complexity
table_complexity_cost(all)
```

**The vaccine.** A ratio below one does not mean the vaccine is free. The
vaccine cuts the median attack rate from `r attack("scenario_00", 1e4)` to
`r attack("scenario_01", 1e4)` at 10,000 agents and from
`r attack("scenario_00", 1e5)` to `r attack("scenario_01", 1e5)` at 100,000,
so engines whose work follows the outbreak (EoN, epydemic, the epiworld
family, and ixa) get faster for that reason alone. Covasim's vectorized daily
update touches every agent regardless, and Starsim draws transmission on
every edge every day, so their times barely move; individual sits in
between. A ratio above one means the feature itself is expensive.

**The outputs.** Both scenarios run the same epidemics, so a ratio above one
is the cost of recording the outputs during the run. The epiworld family
already records all four in every scenario, and Covasim keeps its infection
log in every run, so they pay nothing extra; most engines that had to add
recording absorbed it too. The exceptions are EoN, whose
`return_full_data=True` appends to every node's status history in Python at
every event; individual, whose runner compares full-population bitsets every
day to count transitions; and FRED, which writes a line to its health
records for every exposure and state change. Reading the outputs out of the
engine after the run is not in these times; scenario 02's report shows it.

## Memory

Overall peak resident memory of every replicate (see
[Memory](methods.md#memory)):

```{r}
#| label: memory
#| fig-width: 12
#| fig-height: 9
plot_memory_all(all)
```

Median simulation memory, the extra memory one more replicate needs, in MiB:

```{r}
#| label: simulation-memory
table_by_cell(all, "simulation_mib", 0)
```

At 1,000,000 agents, FRED peaks at
`r mib(med("FRED", "scenario_03", 1e6, "peak_mib"))` MiB, epydemic at
`r mib(med("epydemic", "scenario_03", 1e6, "peak_mib"))`, and EoN at
`r mib(med("EoN", "scenario_03", 1e6, "peak_mib"))`, against
`r mib(med("epiworld", "scenario_03", 1e6, "peak_mib"))` for the C++
epiworld runner, `r mib(med("ixa", "scenario_03", 1e6, "peak_mib"))` for
ixa, and `r mib(med("individual", "scenario_03", 1e6, "peak_mib"))` for
individual. At small sizes the runtime itself can dominate: Agents.jl uses
`r mib(med("Agents.jl", "scenario_00", 1e4, "rss_baseline_bytes") / 2^20)` MiB
before it reads the edge list. At large sizes, most of EoN's memory is its
NetworkX graph (model footprint), epydemic adds about as much again during
the run, and most of ixa's and Starsim's is their per-replicate rebuild
(simulation memory).

## Epidemiological agreement

Median final attack rate by engine, with each engine's difference from
epiworldR's in parentheses:

```{r}
#| label: agreement
table_agreement(all)
```

In scenarios 00 to 03, every engine's median stays within about 0.01 of
epiworldR's. Scenario 04 is uncalibrated (see its report): Covasim sits
`r sprintf("%.3f", med("covasim", "scenario_04", 165865, "attack_rate") - med("epiworldR", "scenario_04", 165865, "attack_rate"))`
above, and every other engine within about 0.01. Each report also shows the peak hospitalization load, and scenario 02's
checks every engine's outputs against its own final counts.

## Sensitivity and validity

- **Outbreak size.** Run time depends on the epidemic as well as on the
  population: the vaccine (scenario 01) shrinks the outbreak and speeds up
  the engines whose work follows it, scenario 03 grows the population ten
  times with an outbreak that hardly grows, and scenario 04's outbreak
  reaches most of its population. Ranking engines by one scenario alone
  would mislead.
- **Concurrency.** Running replicates side by side changes timings unevenly
  across engines, so every published run is sequential
  ([Execution protocol](methods.md#execution-protocol)).
- **Environment.** Every scenario's replicates ran in one recorded
  environment ([Environment](methods.md#environment)); a report warns when
  its replicates mix environments.
- **Validity checks.** Every runner checks that its five compartments sum
  to the population, and scenario 02 checks the outputs against the final
  counts in every run.
- **Not yet covered.** An alternate transmission intensity or outbreak size,
  and prespecified equivalence thresholds for the epidemiological outcomes
  ([#10](https://github.com/UofUEpiBio/epiworld-benchmark/issues/10)).

## Implementation size

Model lines of every engine in every scenario (see
[Implementation size](methods.md#implementation-size)). Scenarios 03 and 04
reuse scenario 00's runners.

```{r}
#| label: code-size
table_code_size(all)
```

## Completion

Completed replicates against each scenario's design, and the failures each
scenario's latest run recorded. Every engine expresses and completes every
scenario.

```{r}
#| label: completion
table_completion(all)
```
